---
title: "10_Emission_Analysis"
output: html_document
date: "2025-09-26"
---
## 18) Link emissions to productivity data 

here we will have the same issue as with feed intake data since the Alevel in the data tab dont always latch the ones in the diet tab 

```{r}
library(dplyr)
library(ERAg)
library(stringr)
library(knitr)
library(dplyr)
library(DT)
library(readxl)
library(tidyr)
library(arrow)
library(ggplot2)
library(data.table)
library(purrr)
library(janitor)
library(ggalt)
library(dplyr)
library(ggplot2)
library(readr)
library(stringr)
library(forcats)

GE_combined      <- read.csv("~/Emission_Analysis/data/GE_combined.csv")          
yield_data       <- read.csv("~/Emission_Analysis/data/yield_data.csv")
weights_unique   <- read.csv("~/Emission_Analysis/data/weights_unique.csv")
practice_tbl     <- read.csv("~/Emission_Analysis/data/practice_tbl.csv")
practice_pairwise_tbl     <- read.csv("~/Emission_Analysis/data/practice_trt_tbl.csv")
merged_metadata <- readRDS("~/Emission_Analysis/data/merged_metadata_nutrition_inference.rds")
classif_ing_simple <- read.csv("~/Emission_Analysis/data/classif_ing_simple.csv")
diet_percentages <- read.csv("~/Emission_Analysis/data/diet_percentages_classif.csv")
diet_classif <- read.csv("~/Emission_Analysis/data/pairwise_diet_shift_tbl.csv")


roman_to_name <- function(x) dplyr::recode(
  x,
  "I"   = "High-starch",
  "II"  = "High-fat",
  "III" = "Protein",
  "IV"  = "Good forage",
  "V"   = "Poor forage",
  "VI"  = "Mineral/Vitamin",
  .default = x,
  .missing = NA_character_
)

# Update the ingredient database
classif_ing_simple <- classif_ing_simple %>%
  dplyr::mutate(
    Ingredient_Category = roman_to_name(Ingredient_Category))
    # If you also keep a diet-level tag, uncomment:
    # , Diet_Category = roman_to_name(Diet_Category)
  

```


### to be removed papers with problems 

```{r}
#EM1047 too high intake not possible 

GE_combined<-GE_combined %>%
  filter(B.Code!="EM1047")

#BO1076 loo at practice tbl becaus esays no change but ther eis 

```


### 18.1) Merge yield and emission data and number of papers 

```{r}

# Step 1: Make sure the key columns match (D.Item and A.Level.Name)
GE_combined_for_merge <- GE_combined %>%
  filter(!is.na(CH4_kg_year))%>%
  rename(A.Level.Name = D.Item)%>%
  select(-"ED.Mean.T")


yield_data <- yield_data %>%
  filter(!is.na(ED.Mean.T))%>%
  filter(Out.Subind!="Meat Yield-Final Body Weight")


# Step 2: Merge yield and emission data
yield_with_emissions <- GE_combined_for_merge %>%
  left_join(
    yield_data,
    by = c("B.Code", "A.Level.Name")
  )%>%
    filter(!is.na(ED.Mean.T))%>%
  filter(!is.na(CH4_kg_year))

  

# distinct B.Code per product
bcode_per_product <- yield_with_emissions %>%          # or yield_data
  distinct(P.Product, B.Code) %>%                      # keep one row per pair
  count(P.Product, name = "n_BCodes")                  # tally them

bcode_per_product

practice_tbl<-practice_tbl%>%
  filter(B.Code %in% yield_with_emissions$B.Code)

length(unique(practice_tbl$B.Code))

```


primary data exploration
```{r}
# ─────────────────────────────────────────────────────────────
# Libraries
# ─────────────────────────────────────────────────────────────
# ─────────────────────────────────────────────────────────────
# Clean and merge: diet_percentages + ingredient classification
# ─────────────────────────────────────────────────────────────
diet_percentages <- diet_percentages %>%
  mutate(D.Item_clean = tolower(str_trim(D.Item)))

classif_ing_simple <- classif_ing_simple %>%
  mutate(D.Item_clean = tolower(str_trim(D.Item)))

diet_pct_classif <- diet_percentages %>%
  left_join(classif_ing_simple %>% select(D.Item_clean, Ingredient_Category),
            by = "D.Item_clean") %>%
  filter(!is.na(Ingredient_Category))

# ─────────────────────────────────────────────────────────────
# FIGURE 1: Barplot of Contribution by Ingredient Category
# ─────────────────────────────────────────────────────────────
diet_pct_classif %>%
  group_by(Ingredient_Category) %>%
  summarise(TotalPct = sum(Percentage_of_Diet, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(Ingredient_Category, TotalPct), y = TotalPct)) +
  geom_col(fill = "forestgreen") +
  coord_flip() +
  labs(
    title = "Contribution of Ingredient Categories to Livestock Diets",
    x = "Ingredient Category",
    y = "Cumulative Share of Diet (%)"
  ) +
  theme_minimal()


# Compute count of unique B.Code per Practice
practice_counts <- practice_tbl %>%
  filter(Practice %in% c("Addition", "Substitution", "No change")) %>%
  group_by(Practice) %>%
  summarise(N_cases = n_distinct(B.Code)) %>%
  ungroup()

# Join back to the main table for labeling inside each facet
practice_labeled <- practice_tbl %>%
  filter(Practice %in% c("Addition", "Substitution", "No change")) %>%
  left_join(practice_counts, by = "Practice")

# Base plot with counts annotated in facet titles
ggplot(practice_labeled, aes(x = Magnitude_pct, fill = Practice)) +
  geom_histogram(binwidth = 2, color = "white") +
  facet_wrap(~ Practice + N_cases, scales = "free_y",
             labeller = labeller(.cols = function(practice) {
               paste0(practice, "\n(n = ", practice_counts$N_cases[match(practice, practice_counts$Practice)], ")")
             })) +
  labs(
    title = "Magnitude of Diet Change by Practice Type",
    x = "Change in % of Diet",
    y = "Number of Cases",
    fill = "Practice"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "none"
  )


# ─────────────────────────────────────────────────────────────
# FIGURE 3: Top Substitution Pairs
# ─────────────────────────────────────────────────────────────
practice_tbl %>%
  filter(Practice == "Substitution", Substitution_pairs_detail != "") %>%
  count(Substitution_pairs_detail, sort = TRUE) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(Substitution_pairs_detail, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Substitution Category Pairs",
    x = "From → To",
    y = "Frequency"
  ) +
  theme_minimal()

# ─────────────────────────────────────────────────────────────
# FIGURE 4: GE vs CH4 Emissions
# ─────────────────────────────────────────────────────────────
GE_combined %>%
  filter(!is.na(CH4_kg_year), !is.na(GE_daily)) %>%
  ggplot(aes(x = GE_daily, y = CH4_kg_year)) +
  geom_point(alpha = 0.6, color = "gray30") +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  labs(
    title = "Methane Emissions vs Gross Energy Intake",
    x = "Gross Energy (MJ/day)",
    y = "Methane (kg CH₄ / year)"
  ) +
  theme_minimal()



```
## Emission in function of the highest cat in each diet 

```{r}
emission_ing_classif<-diet_pct_classif%>%
  left_join(GE_combined_for_merge,by=c("B.Code","A.Level.Name"))%>%
  filter(!is.na(CH4_kg_year))

#number of unique codes 
emission_ing_classif %>%
  group_by(P.Product) %>%
  summarise(n_unique_B = n_distinct(B.Code), .groups = "drop") %>%
  arrange(desc(n_unique_B))

```


```{r}


df <- emission_ing_classif

# 1) Dominant ingredient per diet (max % of diet), keep Product
dom <- df %>%
  group_by(A.Level.Name) %>%
  slice_max(Percentage_of_Diet, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(A.Level.Name, P.Product, dom_category = Ingredient_Category)

# 2) One CH4 value per diet
ch4 <- df %>%
  group_by(A.Level.Name) %>%
  summarise(CH4_kg_year = mean(CH4_kg_year, na.rm = TRUE), .groups = "drop")

# 3) Join and summarise BY PRODUCT × DOMINANT INGREDIENT
summary_by_prod_dom <- dom %>%
  left_join(ch4, by = "A.Level.Name") %>%
  group_by(P.Product, dom_category) %>%
  summarise(
    n_diets    = n(),
    mean_CH4   = mean(CH4_kg_year, na.rm = TRUE),
    median_CH4 = median(CH4_kg_year, na.rm = TRUE),
    sd_CH4     = sd(CH4_kg_year, na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  arrange(P.Product, desc(n_diets))

summary_by_prod_dom

```

```{r}
plot_df <- dom %>%
  left_join(ch4, by = "A.Level.Name") %>%
  filter(CH4_kg_year < 100)   # optional outlier trim; remove if not desired
# get common y-limits across all products (after any filtering you apply)
y_lims <- range(plot_df$CH4_kg_year, na.rm = TRUE)

ggplot(plot_df, aes(x = dom_category, y = CH4_kg_year, fill = dom_category)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.5) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.5, size = 1.4, show.legend = FALSE) +
  facet_wrap(~ P.Product, scales = "fixed") +              # <- same y for all facets
  scale_y_continuous(limits = y_lims, expand = expansion(mult = c(0.02, 0.05))) +
  labs(x = NULL, y = "CH₄ (kg / animal / year)",
       title = "Enteric CH₄ by dominant ingredient category within each product") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 25, hjust = 1))



```

```{r}
suppressPackageStartupMessages({
  library(dplyr); library(stringr); library(tidyr); library(forcats); library(ggplot2)
})

# ---------- helpers ----------
str_clean <- function(x) stringr::str_squish(as.character(x))
has <- function(x, pat) stringr::str_detect(stringr::str_to_lower(coalesce(x, "")),
                                            stringr::str_to_lower(pat))
first_non_na <- function(x){ x <- x[!is.na(x)]; if (length(x)==0) NA else x[1] }

# ---------- 1) CH4 per arm + species from emissions ----------
em_dat <- emission_ing_classif %>%
  mutate(B.Code = str_clean(B.Code), A.Level.Name = str_clean(A.Level.Name)) %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    CH4_kg_year = mean(CH4_kg_year, na.rm = TRUE),
    P.Product   = dplyr::first(na.omit(P.Product)),
    .groups = "drop"
  )

# ---------- 2) Practice/Subpractice map (include controls) ----------
practice_tbl <- practice_tbl %>%
  mutate(B.Code = str_clean(B.Code), Control = str_clean(Control), Trt = str_clean(Trt))

map_trt <- practice_tbl %>%
  transmute(B.Code, A.Level.Name = Trt, Practice, Subpractice, Magnitude_pct)

map_ctrl <- practice_tbl %>%
  distinct(B.Code, A.Level.Name = Control) %>%
  mutate(Practice="Control", Subpractice="", Magnitude_pct = 0)

practice_map <- bind_rows(map_trt, map_ctrl) %>%
  distinct(B.Code, A.Level.Name, .keep_all = TRUE) %>%
  # ---- collapse multi-category Additions into "Mixed" ----
  mutate(Subpractice = if_else(Practice=="Addition" & str_detect(Subpractice, ","),
                               "Mixed", Subpractice))

# ---------- 3) Choose one yield per arm ----------
yld <- yield_with_emissions %>%
  mutate(B.Code = str_clean(B.Code), A.Level.Name = str_clean(A.Level.Name)) %>%
  group_by(B.Code, A.Level.Name) %>%
  mutate(is_yield = str_detect(str_to_lower(coalesce(Out.Subind, "")), "yield")) %>%
  arrange(B.Code, A.Level.Name, desc(is_yield), desc(!is.na(ED.Mean.T))) %>%
  summarise(
    Yield   = first_non_na(ED.Mean.T),
    Product = first_non_na(Out.Subind),
    Units   = first_non_na(Out.Unit),
    .groups = "drop"
  )

# ---------- 4) Final joined table (keeps controls) ----------
final_df <- em_dat %>%
  left_join(practice_map, by = c("B.Code", "A.Level.Name")) %>%
  left_join(yld,         by = c("B.Code", "A.Level.Name")) %>%
  transmute(
    B.Code, A.Level.Name,
    Practice    = coalesce(Practice, "Unknown"),
    Subpractice = coalesce(Subpractice, ""),
    Magnitude_pct,
    CH4_kg_year,
    Yield, Product, Units,
    P.Product
  ) %>%
  filter(Practice != "Unknown") %>%
  drop_na(CH4_kg_year, Yield, Units, P.Product) %>%
  arrange(B.Code, A.Level.Name)%>%
  filter(Practice!="No change")

```

```{r}
# ---------- 5) Compute EI (g CH4 / kg product) ----------
final_with_ei_all <- final_df %>%
  mutate(
    days_per_year = case_when(
      has(Units, "/day")  ~ 365,
      has(Units, "/year") ~ 1,
      TRUE ~ NA_real_
    ),
    yield_kg = case_when(
      has(Units, "kg") ~ Yield,
      has(Units, "g")  ~ Yield/1000,
      has(Units, "l") & has(Product, "milk") ~ Yield * 1.03,  # L→kg for milk
      TRUE ~ NA_real_
    ),
    annual_yield_kg = yield_kg * days_per_year,
    EI_g_per_kg     = (CH4_kg_year * 1000) / annual_yield_kg
  ) %>%
  filter(is.finite(EI_g_per_kg), EI_g_per_kg > 0, is.finite(annual_yield_kg))

# ---------- 6) Deltas vs control (treatment–control within B.Code) ----------
ctrl_map <- practice_tbl %>% distinct(B.Code, Control) %>% rename(Control_Arm = Control)

ctrl_metrics <- final_with_ei_all %>%
  select(B.Code, A.Level.Name, EI_ctrl = EI_g_per_kg,
         CH4_ctrl = CH4_kg_year, Y_ctrl = annual_yield_kg) %>%
  rename(Control_Arm = A.Level.Name)

final_with_deltas <- practice_tbl %>%
  select(B.Code, Control_Arm = Control, A.Level.Name = Trt, Practice, Subpractice, Magnitude_pct) %>%
  # align "Mixed" rule for Additions here too
  mutate(Subpractice = if_else(Practice=="Addition" & str_detect(Subpractice, ","),
                               "Mixed", Subpractice)) %>%
  inner_join(final_with_ei_all, by = c("B.Code","A.Level.Name")) %>%
  inner_join(ctrl_metrics,      by = c("B.Code","Control_Arm")) %>%
  mutate(
    Delta_EI_pct       = 100 * (EI_g_per_kg / EI_ctrl - 1),
    Delta_EI_g_per_kg  = EI_g_per_kg - EI_ctrl,
    Delta_CH4_pct      = 100 * (CH4_kg_year / CH4_ctrl - 1),
    Delta_Y_pct        = 100 * (annual_yield_kg / Y_ctrl - 1))


    
  final_with_deltas <- final_with_deltas %>%
  dplyr::mutate(
    Practice      = dplyr::coalesce(Practice.x, Practice.y),
    Subpractice   = dplyr::coalesce(Subpractice.x, Subpractice.y),
    Magnitude_pct = dplyr::coalesce(Magnitude_pct.x, Magnitude_pct.y)
  ) %>%
  dplyr::select(
    B.Code, Control_Arm, A.Level.Name,
    Practice, Subpractice, Magnitude_pct,
    CH4_kg_year, Yield, Product, Units, P.Product,
    days_per_year, yield_kg, annual_yield_kg, EI_g_per_kg,
    EI_ctrl, CH4_ctrl, Y_ctrl,
    Delta_EI_pct, Delta_EI_g_per_kg, Delta_CH4_pct, Delta_Y_pct
  ) %>%
  dplyr::distinct()

# (optional) keep only treatment rows (i.e., exclude the control arm itself)
final_with_deltas <- final_with_deltas %>%
  dplyr::filter(A.Level.Name != Control_Arm)

  

```


```{r}
# ---- A) EI by Practice (treatments only; point overlay + median diamond) ----
ei_practice_df <- final_with_ei_all %>% filter(!Practice %in% c("Control","No change"))

ggplot(ei_practice_df, aes(x = Practice, y = EI_g_per_kg, fill = Practice)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.8, fill = "white", color = "black") +
  geom_jitter(width = 0.15, alpha = 0.55, size = 1.4, show.legend = FALSE) +
  labs(x = NULL, y = expression("Emission intensity (g CH"[4]*"/ kg product)"),
       title = "Emission intensity by practice") +
  theme_minimal(12) + theme(legend.position = "none")

# ---- B) ΔEI (%) by Practice (treatments vs control) ----
ggplot(final_with_deltas, aes(x = Practice, y = Delta_EI_pct, fill = Practice)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, alpha = 0.55, size = 1.3, show.legend = FALSE) +
  labs(x = NULL, y = "Δ Emission intensity (% vs control)",
       title = "Δ Emission intensity by practice") +
  theme_minimal(12) + theme(legend.position = "none")

```

# for pairwise tbl
```{r}
# =====================================================================
#      PAIRWISE EMISSION–YIELD–PRACTICE WORKFLOW  (from scratch)
# =====================================================================

suppressPackageStartupMessages({
  library(dplyr); library(stringr); library(tidyr); library(forcats); library(ggplot2)
})

# ---------- helpers ----------
str_clean <- function(x) stringr::str_squish(as.character(x))
has <- function(x, pat) stringr::str_detect(stringr::str_to_lower(coalesce(x, "")),
                                            stringr::str_to_lower(pat))
first_non_na <- function(x){ x <- x[!is.na(x)]; if (length(x)==0) NA else x[1] }

# =====================================================================
# 1️⃣ Emission data per diet arm
# =====================================================================
em_dat <- emission_ing_classif %>%
  mutate(
    B.Code = str_clean(B.Code),
    A.Level.Name = str_clean(A.Level.Name)
  ) %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    CH4_kg_year = mean(CH4_kg_year, na.rm = TRUE),
    P.Product   = first_non_na(P.Product),
    .groups = "drop"
  )

# =====================================================================
# 2️⃣ Map practice / subpractice from pairwise table
# =====================================================================
pairwise_practice <- practice_pairwise_tbl %>%
  mutate(
    B.Code = str_clean(B.Code),
    A = str_clean(A),
    B = str_clean(B),
    Subpractice = if_else(
      Practice == "Addition" & str_detect(Subpractice, ","),
      "Mixed",
      Subpractice
    )
  )

# We'll create a long "arm-level" map for both A and B diets
map_A <- pairwise_practice %>%
  transmute(
    B.Code,
    A.Level.Name = A,
    Practice,
    Subpractice,
    Magnitude_pct,
    pair_label = paste(A, "vs", B),
    arm_role = "A"
  )

map_B <- pairwise_practice %>%
  transmute(
    B.Code,
    A.Level.Name = B,
    Practice,
    Subpractice,
    Magnitude_pct,
    pair_label = paste(A, "vs", B),
    arm_role = "B"
  )

practice_map <- bind_rows(map_A, map_B) %>%
  distinct(B.Code, A.Level.Name, pair_label, arm_role, .keep_all = TRUE)

# =====================================================================
# 3️⃣ Yield data per diet arm
# =====================================================================
yld <- yield_with_emissions %>%
  mutate(
    B.Code = str_clean(B.Code),
    A.Level.Name = str_clean(A.Level.Name)
  ) %>%
  group_by(B.Code, A.Level.Name) %>%
  mutate(is_yield = str_detect(str_to_lower(coalesce(Out.Subind, "")), "yield")) %>%
  arrange(B.Code, A.Level.Name, desc(is_yield), desc(!is.na(ED.Mean.T))) %>%
  summarise(
    Yield   = first_non_na(ED.Mean.T),
    Product = first_non_na(Out.Subind),
    Units   = first_non_na(Out.Unit),
    .groups = "drop"
  )

# =====================================================================
# 4️⃣ Merge: Emissions + Yield + Practice
# =====================================================================
final_df <- em_dat %>%
  left_join(practice_map, by = c("B.Code", "A.Level.Name")) %>%
  left_join(yld, by = c("B.Code", "A.Level.Name")) %>%
  transmute(
    B.Code, A.Level.Name,
    Practice    = coalesce(Practice, "Unknown"),
    Subpractice = coalesce(Subpractice, ""),
    Magnitude_pct,
    CH4_kg_year, Yield, Product, Units, P.Product,
    pair_label, arm_role
  ) %>%
  filter(Practice != "Unknown") %>%
  drop_na(CH4_kg_year, Yield, Units, P.Product) %>%
  arrange(B.Code, A.Level.Name)

# =====================================================================
# 5️⃣ Compute EI (Emission Intensity)
# =====================================================================
final_with_ei_all_pairwise <- final_df %>%
  mutate(
    days_per_year = case_when(
      has(Units, "/day")  ~ 365,
      has(Units, "/year") ~ 1,
      TRUE ~ NA_real_
    ),
    yield_kg = case_when(
      has(Units, "kg") ~ Yield,
      has(Units, "g")  ~ Yield / 1000,
      has(Units, "l") & has(Product, "milk") ~ Yield * 1.03,  # convert milk L→kg
      TRUE ~ NA_real_
    ),
    annual_yield_kg = yield_kg * days_per_year,
    EI_g_per_kg     = (CH4_kg_year * 1000) / annual_yield_kg
  ) %>%
  filter(is.finite(EI_g_per_kg), EI_g_per_kg > 0, is.finite(annual_yield_kg))

# =====================================================================
# 6️⃣ Compute pairwise deltas (A vs B within each B.Code)
# =====================================================================
# Join back A and B values from the same pair
A_info <- final_with_ei_all_pairwise %>%
  filter(arm_role == "A") %>%
  rename(
    CH4_A = CH4_kg_year,
    EI_A = EI_g_per_kg,
    Y_A = annual_yield_kg,
    Product_A = Product
  ) %>%
  select(B.Code, pair_label, Practice, Subpractice, Magnitude_pct,
         CH4_A, EI_A, Y_A, Product_A)

B_info <- final_with_ei_all_pairwise %>%
  filter(arm_role == "B") %>%
  rename(
    CH4_B = CH4_kg_year,
    EI_B = EI_g_per_kg,
    Y_B = annual_yield_kg,
    Product_B = Product
  ) %>%
  select(B.Code, pair_label, Practice, Subpractice, Magnitude_pct,
         CH4_B, EI_B, Y_B, Product_B)

final_with_deltas_pairwise <- inner_join(A_info, B_info, by = c("B.Code", "pair_label", "Practice", "Subpractice", "Magnitude_pct")) %>%
  mutate(
    Delta_EI_pct       = 100 * (EI_B / EI_A - 1),
    Delta_EI_g_per_kg  = EI_B - EI_A,
    Delta_CH4_pct      = 100 * (CH4_B / CH4_A - 1),
    Delta_Y_pct        = 100 * (Y_B / Y_A - 1)
  ) %>%
  filter(is.finite(Delta_EI_pct))

# =====================================================================
# 7️⃣ Visualization
# =====================================================================

# ---- EI by Practice ----
ggplot(final_with_ei_all_pairwise %>% filter(arm_role == "B"), 
       aes(x = Practice, y = EI_g_per_kg, fill = Practice)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.2, show.legend = FALSE) +
  labs(
    x = NULL,
    y = expression("Emission intensity (g CH"[4]*"/ kg product)"),
    title = "Emission intensity by practice (pairwise comparison)"
  ) +
  theme_minimal(12) + theme(legend.position = "none")

# ---- ΔEI (%) by Practice ----
ggplot(final_with_deltas_pairwise, aes(x = Practice, y = Delta_EI_pct, fill = Practice)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.3, show.legend = FALSE) +
  labs(
    x = NULL,
    y = expression(Delta*" Emission intensity (% vs paired arm)"),
    title = "Δ Emission intensity by practice (pairwise A–B)"
  ) +
  theme_minimal(12) + theme(legend.position = "none")

```

```{r}
# ---- C) EI by Subpractice (with Ns in labels; reorder by median) ----
labs_sub <- final_with_ei_all %>%
  count(Subpractice, name = "n") %>%
  mutate(Subpractice_lab = paste0(Subpractice, " (n=", n, ")"))

plot_ei_sub <- final_with_ei_all %>%
  left_join(labs_sub, by = "Subpractice") %>%
  mutate(Subpractice_lab = fct_reorder(Subpractice_lab, EI_g_per_kg, median, na.rm = TRUE))

ggplot(plot_ei_sub,
       aes(x = Subpractice_lab, y = EI_g_per_kg, fill = Subpractice)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, alpha = 0.55, size = 1.2, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = expression("Emission intensity (g CH"[4]*"/ kg product)"),
       title = "Emission intensity by subpractice (dominant diet lever)") +
  theme_minimal(12) + theme(legend.position = "none")

# ---- D) ΔEI (%) by Subpractice (with points + median; Ns; reordered) ----
labs_sub_delta <- final_with_deltas %>%
  count(Subpractice, name = "n") %>%
  mutate(Subpractice_lab = paste0(Subpractice, " (n=", n, ")"))

plot_delta_sub <- final_with_deltas %>%
  left_join(labs_sub_delta, by = "Subpractice") %>%
  mutate(Subpractice_lab = fct_reorder(Subpractice_lab, Delta_EI_pct, median, na.rm = TRUE))

ggplot(plot_delta_sub,
       aes(x = Subpractice_lab, y = Delta_EI_pct, fill = Subpractice)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.55, size = 1.4, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Δ Emission intensity (% vs control)",
       title = "Δ Emission intensity by subpractice (dominant diet lever)") +
  theme_minimal(12) + theme(legend.position = "none")

plot_delta_sub <- final_with_deltas %>%
  filter(Practice=="Addition")%>%
  left_join(labs_sub_delta, by = "Subpractice") %>%
  mutate(Subpractice_lab = fct_reorder(Subpractice_lab, Delta_EI_pct, median, na.rm = TRUE))

ggplot(plot_delta_sub,
       aes(x = Subpractice_lab, y = Delta_EI_pct, fill = Subpractice)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.55, size = 1.4, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Δ Emission intensity (% vs control)",
       title = "Δ Emission intensity by subpractice (dominant diet lever)") +
  theme_minimal(12) + theme(legend.position = "none")


ggplot(plot_delta_sub,
       aes(x = Subpractice_lab, y = Delta_Y_pct, fill = Subpractice)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.55, size = 1.4, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Δ yield intensity (% vs control)",
       title = "Δ yield  by subpractice (dominant diet lever)") +
  theme_minimal(12) + theme(legend.position = "none")

```
 interpretation : 
 
 Across the compiled dataset, emission intensity (EI) responses to dietary interventions showed consistent but nuanced patterns reflecting the interaction between productivity gains and methane output. When expressed in absolute terms (g CH₄ kg⁻¹ product), diets dominated by high-quality forages and moderate lipid inclusion exhibited the lowest emission intensities (Figure X). These diets typically enhanced feed digestibility and nutrient use efficiency, leading to reduced methane loss per unit of product. In contrast, interventions based on high-starch or protein supplementation displayed wider variability, indicating that improvements in productivity did not always translate proportionally into lower emissions. Substitution practices combining changes in multiple diet components also showed large dispersion, highlighting the heterogeneous nature of feeding experiments across African livestock systems.

When examined as relative change compared to control diets (Δ EI, %), most dietary practices clustered around zero or below (Figure X). Values near zero are particularly meaningful in this context: they can represent either a true lack of response—where methane emissions and yield changed in similar proportions—or a compensatory effect, in which productivity improvements offset higher methane output, resulting in no net change in EI. This neutral outcome does not necessarily imply inefficacy but rather that interventions increased both feed intake and animal performance proportionally. Indeed, several “Good forage” and “Protein” treatments produced modest methane increases alongside higher yields, leading to stable emission intensities relative to controls. Conversely, negative Δ EI values (reductions) occurred where nutrient density or digestibility improvements led to greater productivity gains than methane increases.

The number of data points varied substantially among subpractices, influencing the strength and interpretation of observed trends. For instance, Good forage additions (n = 18) and Protein supplementation (n = 6) were well represented and yielded consistent reductions or neutral changes in EI, while categories such as High-starch (n = 2) and Poor forage (n = 1) were based on few observations and should be interpreted cautiously. Mixed interventions (n = 5), often involving multiple nutrient adjustments, produced the largest spread in Δ EI, reflecting both synergistic and antagonistic effects among combined practices. The broad variability across categories underscores the diversity of study conditions, ranging from controlled feeding trials to field-based studies with variable diet formulation accuracy.

Productivity responses followed a generally positive trend across interventions (Figure X). Most practices increased yield relative to controls, with mixed or multi-component strategies producing the largest median improvements, often exceeding 50%. High-fat and protein supplementation resulted in more moderate gains, typically 5–15%. These results confirm that interventions improving diet quality tend to enhance animal performance while either reducing or maintaining emission intensity per unit of product.

Overall, the data indicate that dietary interventions targeting improved feed quality and nutrient balance yield dual benefits for productivity and methane mitigation in African livestock systems. The results also highlight that stable or “zero-change” emission intensity outcomes can represent balanced improvements in both emissions and yield—an important nuance in interpreting mitigation efficacy. Despite the inherent variability and uneven data coverage across practices, the overall direction of change supports feed quality enhancement—through improved forages, protein supplementation, and moderate fat inclusion—as a robust strategy for reducing methane emission intensity and increasing productivity across diverse African ruminant production systems.
### for pairwise tbl 
```{r}
# ---- C) EI by Subpractice & Practice (pairwise) ----
labs_sub_pw <- final_with_ei_all_pairwise %>%
  count(Practice, Subpractice, name = "n") %>%
  mutate(Subpractice_lab = paste0(Subpractice, " (n=", n, ")"))

plot_ei_sub_pw <- final_with_ei_all_pairwise %>%
  left_join(labs_sub_pw, by = c("Practice", "Subpractice")) %>%
  mutate(Subpractice_lab = fct_reorder(Subpractice_lab, EI_g_per_kg, median, na.rm = TRUE))

ggplot(plot_ei_sub_pw,
       aes(x = Subpractice_lab, y = EI_g_per_kg, fill = Practice)) +
  geom_boxplot(alpha = 0.65, outlier.alpha = 0.3, color = "gray35") +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.4, fill = "white", color = "black") +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.2, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = expression("Emission intensity (g CH"[4]*"/ kg product)"),
       title = "Emission intensity by Subpractice (pairwise data)") +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c("Addition" = "#1b9e77",
                               "Substitution" = "#7570b3",
                               "Reduction" = "#d95f02",
                               "No change" = "gray70")) +
  theme(legend.title = element_blank(),
        legend.position = "top")

# ---- D) ΔEI (%) by Subpractice & Practice (pairwise) ----
labs_sub_delta_pw <- final_with_deltas_pairwise %>%
  count(Practice, Subpractice, name = "n") %>%
  mutate(Subpractice_lab = paste0(Subpractice, " (n=", n, ")"))

plot_delta_sub_pw <- final_with_deltas_pairwise %>%
  left_join(labs_sub_delta_pw, by = c("Practice", "Subpractice")) %>%
  mutate(Subpractice_lab = fct_reorder(Subpractice_lab, Delta_EI_pct, median, na.rm = TRUE))

ggplot(plot_delta_sub_pw,
       aes(x = Subpractice_lab, y = Delta_EI_pct, fill = Practice)) +
  geom_hline(yintercept = 0, linetype = 2, color = "gray50") +
  geom_boxplot(alpha = 0.65, outlier.alpha = 0.3, color = "gray35") +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.4, fill = "white", color = "black") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.55, size = 1.3, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Δ Emission intensity (% vs paired arm)",
       title = "Δ Emission intensity by Subpractice (pairwise data)") +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c("Addition" = "#1b9e77",
                               "Substitution" = "#7570b3",
                               "Reduction" = "#d95f02",
                               "No change" = "gray70")) +
  theme(legend.title = element_blank(),
        legend.position = "top")

# ---- E) ΔYield (%) by Subpractice & Practice (pairwise) ----
plot_delta_yield_pw <- final_with_deltas_pairwise %>%
  left_join(labs_sub_delta_pw, by = c("Practice", "Subpractice")) %>%
  mutate(Subpractice_lab = fct_reorder(Subpractice_lab, Delta_Y_pct, median, na.rm = TRUE))

ggplot(plot_delta_yield_pw,
       aes(x = Subpractice_lab, y = Delta_Y_pct, fill = Practice)) +
  geom_hline(yintercept = 0, linetype = 2, color = "gray50") +
  geom_boxplot(alpha = 0.65, outlier.alpha = 0.3, color = "gray35") +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.4, fill = "white", color = "black") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.55, size = 1.3, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Δ Yield (% vs paired arm)",
       title = "Δ Yield by Subpractice (pairwise data)") +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c("Addition" = "#1b9e77",
                               "Substitution" = "#7570b3")) +
  theme(legend.title = element_blank(),
        legend.position = "top")

```


### 18.2) Plot the data
```{r}


# Filter meat yield data and drop missing values
yield_meat <- yield_with_emissions %>%
  filter(Out.Subind == "Meat Yield") %>%
  filter(!is.na(ED.Mean.T) & !is.na(CH4_kg_year) & !is.na(P.Product))

# Get axis limits
x_range <- range(yield_meat$ED.Mean.T, na.rm = TRUE)
y_range <- range(yield_meat$CH4_kg_year, na.rm = TRUE)

# Plot
ggplot(yield_meat, aes(x = ED.Mean.T, y = CH4_kg_year, color = P.Product)) +
  geom_point(alpha = 0.8, size = 2.5) +
  scale_x_log10(
    limits = x_range,
    breaks = scales::log_breaks(n = 5),
    labels = scales::comma_format()
  ) +
  scale_y_continuous(
    limits = y_range,
    labels = scales::comma_format()
  ) +
  labs(
    title = "CH4 Emissions vs. Meat Yield",
    x = "Meat Yield per Individual (kg)",
    y = "Annual CH4 Emissions (kg)",
    color = "Livestock Species"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

```{r}
# Filter meat yield data and drop missing values
yield_milk <- yield_with_emissions %>%
  filter(Out.Subind == "Milk Yield") %>%
  filter(!is.na(ED.Mean.T) & !is.na(CH4_kg_year) & !is.na(P.Product))

# Get axis limits
x_range <- range(yield_milk$ED.Mean.T, na.rm = TRUE)
y_range <- range(yield_milk$CH4_kg_year, na.rm = TRUE)

# Plot
ggplot(yield_milk, aes(x = ED.Mean.T, y = CH4_kg_year, color = P.Product)) +
  geom_point(alpha = 0.8, size = 2.5) +
  scale_x_log10(
    limits = x_range,
    breaks = scales::log_breaks(n = 5),
    labels = scales::comma_format()
  ) +
  scale_y_continuous(
    limits = y_range,
    labels = scales::comma_format()
  ) +
  labs(
    title = "CH4 Emissions vs. Milk Yield",
    x = "Milk Yield per Individual (kg)",
    y = "Annual CH4 Emissions (kg)",
    color = "Livestock Species"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```


```{r}
# Filter and clean data
emissions_by_species <- yield_with_emissions %>%
  filter(!is.na(CH4_kg_year), !is.na(P.Product))

# Boxplot with transparent points
ggplot(emissions_by_species, aes(x = P.Product, y = CH4_kg_year, fill = P.Product)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4) +  # lighter fill and no outliers
  geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = 2) +  # transparent points
  labs(
    title = "Annual CH4 Emissions by Livestock Species",
    x = "Species",
    y = "Annual CH4 Emissions (kg)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  # remove redundant legend

```


```{r}
# Filter and clean data for Goat and Sheep only
emissions_by_species <- yield_with_emissions %>%
  filter(!is.na(CH4_kg_year), P.Product %in% c("Goat", "Sheep"))

# Boxplot with transparent points for Goat and Sheep
ggplot(emissions_by_species, aes(x = P.Product, y = CH4_kg_year, fill = P.Product)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = 2) +
  labs(
    title = "Annual CH4 Emissions: Goat vs. Sheep",
    x = "Species",
    y = "Annual CH4 Emissions (kg)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```




```{r}
##############################################
##  1.  PREPARE DAILY YIELD (kg animal-1 d-1)
##############################################

## density of milk for L → kg conversion
MILK_DENSITY <- 1.03   # kg L-1

yield_d_clean <- yield_data %>%                      # ← your harmonised table
  filter(Out.Subind %in% c("Milk Yield", "Meat Yield")) %>%
  mutate(
    Yield_kg_day = case_when(
      # ───────────── Milk ─────────────────────────────────────────────
      Out.Subind == "Milk Yield" & Out.Unit == "l/individual/day"  ~ ED.Mean.T * MILK_DENSITY,
      Out.Subind == "Milk Yield" & Out.Unit == "kg/individual/day" ~ ED.Mean.T,
      
      # ───────────── Meat (already kg day-1) ─────────────────────────
      Out.Subind == "Meat Yield" & Out.Unit == "kg/individual/day" ~ ED.Mean.T,
      
      # ───────────── Everything else (drop) ─────────────────────────
      TRUE ~ NA_real_
    )
  ) %>%
  filter(!is.na(Yield_kg_day) & Yield_kg_day > 0) %>%   # keep usable rows only
  select(B.Code, A.Level.Name, Out.Subind, Yield_kg_day)


###########################################################################
##  2.  COMPUTE EMISSION-INTENSITY  (kg CH4 per kg *product*) #############
###########################################################################

ei_prod <- GE_combined %>%                              # absolute CH4 table
  select(B.Code, A.Level.Name = D.Item,
         CH4_kg_year, P.Product,T.Control) %>%
  inner_join(yield_d_clean, by = c("B.Code","A.Level.Name")) %>%
  mutate(
    Yield_kg_year = Yield_kg_day * 365,
    EI_CH4_prod   = CH4_kg_year / Yield_kg_year,
    Product       = if_else(Out.Subind == "Milk Yield", "Milk", "Meat")
  ) %>%
  filter(EI_CH4_prod > 0, EI_CH4_prod < 10)             # drop negatives / crazy outliers

#########################################
##  3.  QUICK DIAGNOSTIC PLOTS ##########
#########################################

# ── A) Box-and-jitter by species & product ───────────────────────────────
ggplot(ei_prod, aes(x = P.Product, y = EI_CH4_prod,
                    fill = P.Product)) +
  geom_boxplot(outlier.shape = NA, alpha = .35) +
  geom_jitter(width = .15, alpha = .65, size = 2) +
  scale_y_log10("Emission-intensity (kg CH₄ kg⁻¹ product)",
                breaks = scales::log_breaks(),
                labels = scales::comma_format()) +
  facet_wrap(~ Product, nrow = 1) +
  labs(title = "Enteric-methane intensity of milk and meat production",
       x = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")


```

## diet shift analysis

```{r}

# =====================================================================
#   DIET SHIFT × EMISSION × YIELD ANALYSIS
#   Goal: combine emission + yield + diet shift classifications
# =====================================================================

suppressPackageStartupMessages({
  library(dplyr); library(stringr); library(tidyr); library(forcats); library(ggplot2)
})

# ---------------------------------------------------------------------
# 0️⃣ Helper functions
# ---------------------------------------------------------------------
str_clean <- function(x) stringr::str_squish(as.character(x))
has <- function(x, pat) stringr::str_detect(stringr::str_to_lower(coalesce(x, "")),
                                            stringr::str_to_lower(pat))
first_non_na <- function(x){ x <- x[!is.na(x)]; if (length(x)==0) NA else x[1] }

# =====================================================================
# 1️⃣ EMISSION DATA PER ARM
# =====================================================================
em_dat <- emission_ing_classif %>%
  mutate(B.Code = str_clean(B.Code), A.Level.Name = str_clean(A.Level.Name)) %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    CH4_kg_year = mean(CH4_kg_year, na.rm = TRUE),
    P.Product   = first_non_na(P.Product),
    .groups = "drop"
  )

# =====================================================================
# 2️⃣ DIET SHIFT CLASSIFICATION MAP
# =====================================================================
diet_shift_tbl <- diet_classif %>%
  mutate(
    B.Code = str_clean(B.Code),
    A = str_clean(A),
    B = str_clean(B),
    Subpractice = if_else(
      Practice == "Diet shift" & str_detect(Subpractice, ","),
      "Mixed",
      Subpractice
    )
  )

# Create long arm-level version (A and B)
map_A <- diet_shift_tbl %>%
  transmute(
    B.Code,
    A.Level.Name = A,
    Practice,
    Subpractice,
    pair_label = paste(A, "vs", B),
    arm_role = "A"
  )

map_B <- diet_shift_tbl %>%
  transmute(
    B.Code,
    A.Level.Name = B,
    Practice,
    Subpractice,
    pair_label = paste(A, "vs", B),
    arm_role = "B"
  )


diet_map <- bind_rows(map_A, map_B) %>%
  distinct(B.Code, A.Level.Name, pair_label, arm_role, .keep_all = TRUE)

# =====================================================================
# 3️⃣ YIELD DATA PER ARM
# =====================================================================
yld <- yield_with_emissions %>%
  mutate(B.Code = str_clean(B.Code), A.Level.Name = str_clean(A.Level.Name)) %>%
  group_by(B.Code, A.Level.Name) %>%
  mutate(is_yield = str_detect(str_to_lower(coalesce(Out.Subind, "")), "yield")) %>%
  arrange(B.Code, A.Level.Name, desc(is_yield), desc(!is.na(ED.Mean.T))) %>%
  summarise(
    Yield   = first_non_na(ED.Mean.T),
    Product = first_non_na(Out.Subind),
    Units   = first_non_na(Out.Unit),
    .groups = "drop"
  )

str_clean <- function(x) {
  x <- tolower(as.character(x))       # lower case
  x <- gsub("\\s+", " ", x)           # collapse multiple spaces
  trimws(x)                           # trim leading/trailing spaces
}

# --- Apply consistent cleaning to join keys ---
em_dat <- em_dat %>%
  mutate(
    B.Code = str_clean(B.Code),
    A.Level.Name = str_clean(A.Level.Name)
  )

yld <- yld %>%
  mutate(
    B.Code = str_clean(B.Code),
    A.Level.Name = str_clean(A.Level.Name)
  )

diet_map <- diet_map %>%
  mutate(
    B.Code = str_clean(B.Code),
    A.Level.Name = str_clean(A.Level.Name)
  )


# =====================================================================
# 4️⃣ Merge emissions + yield + diet shift map
# =====================================================================
final_df <- em_dat %>%
  left_join(diet_map, by = c("B.Code", "A.Level.Name")) %>%
  left_join(yld, by = c("B.Code", "A.Level.Name")) %>%
  transmute(
    B.Code, A.Level.Name,
    Practice    = coalesce(Practice, "Unknown"),
    Subpractice = coalesce(Subpractice, ""),
    CH4_kg_year, Yield, Product, Units, P.Product,
    pair_label, arm_role
  ) %>%
  filter(Practice != "Unknown") %>%
  drop_na(CH4_kg_year, Yield, Units, P.Product) %>%
  arrange(B.Code, A.Level.Name)

# =====================================================================
# 5️⃣ Compute Emission Intensity (EI)
# =====================================================================
final_with_ei_all_diet <- final_df %>%
  mutate(
    days_per_year = case_when(
      has(Units, "/day")  ~ 365,
      has(Units, "/year") ~ 1,
      TRUE ~ NA_real_
    ),
    yield_kg = case_when(
      has(Units, "kg") ~ Yield,
      has(Units, "g")  ~ Yield / 1000,
      has(Units, "l") & has(Product, "milk") ~ Yield * 1.03,
      TRUE ~ NA_real_
    ),
    annual_yield_kg = yield_kg * days_per_year,
    EI_g_per_kg     = (CH4_kg_year * 1000) / annual_yield_kg
  ) %>%
  filter(is.finite(EI_g_per_kg), EI_g_per_kg > 0, is.finite(annual_yield_kg))

# =====================================================================
# 6️⃣ Compute pairwise deltas (A vs B) for each B.Code
# =====================================================================
A_info <- final_with_ei_all_diet %>%
  filter(arm_role == "A") %>%
  rename(
    CH4_A = CH4_kg_year,
    EI_A = EI_g_per_kg,
    Y_A = annual_yield_kg,
    Product_A = Product
  ) %>%
  select(B.Code, pair_label, Practice, Subpractice, CH4_A, EI_A, Y_A, Product_A)

B_info <- final_with_ei_all_diet %>%
  filter(arm_role == "B") %>%
  rename(
    CH4_B = CH4_kg_year,
    EI_B = EI_g_per_kg,
    Y_B = annual_yield_kg,
    Product_B = Product
  ) %>%
  select(B.Code, pair_label, Practice, Subpractice, CH4_B, EI_B, Y_B, Product_B)

final_with_deltas_diet <- inner_join(A_info, B_info, by = c("B.Code", "pair_label", "Practice", "Subpractice")) %>%
  mutate(
    Delta_EI_pct       = 100 * (EI_B / EI_A - 1),
    Delta_EI_g_per_kg  = EI_B - EI_A,
    Delta_CH4_pct      = 100 * (CH4_B / CH4_A - 1),
    Delta_Y_pct        = 100 * (Y_B / Y_A - 1)
  ) %>%
  filter(is.finite(Delta_EI_pct))

# =====================================================================
# 7️⃣ Visualizations
# =====================================================================

# ---- A) EI by Practice (Diet shift vs No change) ----
ggplot(final_with_ei_all_diet %>% filter(arm_role == "B"),
       aes(x = Practice, y = EI_g_per_kg, fill = Practice)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.8, fill = "white", color = "black") +
  geom_jitter(width = 0.15, alpha = 0.55, size = 1.4, show.legend = FALSE) +
  labs(x = NULL, y = expression("Emission intensity (g CH"[4]*"/ kg product)"),
       title = "Emission intensity by diet shift vs no change") +
  theme_minimal(base_size = 12) + theme(legend.position = "none")

# ---- B) ΔEI (%) by Practice ----
ggplot(final_with_deltas_diet, aes(x = Practice, y = Delta_EI_pct, fill = Practice)) +
  geom_hline(yintercept = 0, linetype = 2, color = "gray50") +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, alpha = 0.55, size = 1.3, show.legend = FALSE) +
  labs(x = NULL, y = expression(Delta*" Emission intensity (% vs paired arm)"),
       title = "Δ Emission intensity by diet shift practice") +
  theme_minimal(base_size = 12) + theme(legend.position = "none")

# ---- C) EI by Subpractice (Diet shift only) ----
labs_sub_diet <- final_with_ei_all_diet %>%
  filter(Practice == "Diet shift") %>%
  count(Subpractice, name = "n") %>%
  mutate(Subpractice_lab = paste0(Subpractice, " (n=", n, ")"))

plot_ei_sub_diet <- final_with_ei_all_diet %>%
  filter(Practice == "Diet shift") %>%
  left_join(labs_sub_diet, by = "Subpractice") %>%
  mutate(Subpractice_lab = fct_reorder(Subpractice_lab, EI_g_per_kg, median, na.rm = TRUE))

ggplot(plot_ei_sub_diet,
       aes(x = Subpractice_lab, y = EI_g_per_kg, fill = Subpractice)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, alpha = 0.55, size = 1.2, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = expression("Emission intensity (g CH"[4]*"/ kg product)"),
       title = "Emission intensity by subpractice (diet shift only)") +
  theme_minimal(12) + theme(legend.position = "none")

# ---- D) ΔEI (%) by Subpractice ----
labs_sub_delta_diet <- final_with_deltas_diet %>%
  filter(Practice == "Diet shift") %>%
  count(Subpractice, name = "n") %>%
  mutate(Subpractice_lab = paste0(Subpractice, " (n=", n, ")"))

plot_delta_sub_diet <- final_with_deltas_diet %>%
  filter(Practice == "Diet shift") %>%
  left_join(labs_sub_delta_diet, by = "Subpractice") %>%
  mutate(Subpractice_lab = fct_reorder(Subpractice_lab, Delta_EI_pct, median, na.rm = TRUE))

ggplot(plot_delta_sub_diet,
       aes(x = Subpractice_lab, y = Delta_EI_pct, fill = Subpractice)) +
  geom_hline(yintercept = 0, linetype = 2, color = "gray50") +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.55, size = 1.3, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Δ Emission intensity (% vs paired arm)",
       title = "Δ Emission intensity by subpractice (diet shift only)") +
  theme_minimal(12) + theme(legend.position = "none")

# ---- E) ΔYield (%) by Subpractice ----
ggplot(plot_delta_sub_diet,
       aes(x = Subpractice_lab, y = Delta_Y_pct, fill = Subpractice)) +
  geom_hline(yintercept = 0, linetype = 2, color = "gray50") +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.35) +
  stat_summary(fun = median, geom = "point",
               shape = 23, size = 2.6, fill = "white", color = "black") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.55, size = 1.4, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Δ Yield (% vs paired arm)",
       title = "Δ Yield by subpractice (diet shift only)") +
  theme_minimal(12) + theme(legend.position = "none")



```
## number of papers 

```{r}

# ─────────────────────────────────────────────────────────────


# Helper
norm <- function(x) tolower(trimws(x))

# 1️⃣ Papers with emissions (non-missing CH4)
em_with_ch4 <- GE_combined %>%
  filter(!is.na(CH4_kg_year)) %>%
  distinct(B.Code, P.Product)

cat("\n✅ Papers with CH4 emissions available:\n")
print(head(em_with_ch4, 10))
cat("Total =", n_distinct(em_with_ch4$B.Code), "papers\n")

# 2️⃣ Papers with yield information
yield_with_info <- yield_data %>%
  filter(!is.na(ED.Mean.T)) %>%
  distinct(B.Code, Out.Subind, Out.Unit)

cat("\n✅ Papers with yield data available:\n")
print(head(yield_with_info, 10))
cat("Total =", n_distinct(yield_with_info$B.Code), "papers\n")

# 3️⃣ Papers with ingredient classification (through diet_percentages)
ingredient_papers <- diet_percentages %>%
  mutate(D.Item_clean = norm(D.Item)) %>%
  inner_join(
    classif_ing_simple %>%
      mutate(D.Item_clean = norm(D.Item)) %>%
      distinct(D.Item_clean, Ingredient_Category),
    by = "D.Item_clean"
  ) %>%
  distinct(B.Code, Ingredient_Category)

cat("\n✅ Papers with at least one classified ingredient:\n")
print(head(ingredient_papers, 10))
cat("Total =", n_distinct(ingredient_papers$B.Code), "papers\n")

# 4️⃣ Papers with any practice information
ing_practice_papers <- practice_tbl %>%
  distinct(B.Code, Practice)

cat("\n✅ Papers with practice-level data:\n")
print(head(ing_practice_papers, 10))
cat("Total =", n_distinct(ing_practice_papers$B.Code), "papers\n")

# 5️⃣ Combine all coverage conditions
ingredient_with_yield <- ingredient_papers %>%
  inner_join(ing_practice_papers, by = "B.Code") %>%
  inner_join(em_with_ch4, by = "B.Code") %>%
  inner_join(yield_with_info, by = "B.Code") %>%
  distinct(B.Code, P.Product, Ingredient_Category, Practice)

cat("\n✅ Combined dataset (ingredient + practice + emission + yield):\n")
print(head(ingredient_with_yield, 15))
cat("Total =", n_distinct(ingredient_with_yield$B.Code), "papers\n")

# 6️⃣ Summarize by product
ingredient_summary <- ingredient_with_yield %>%
  filter(Practice!="No change")%>%
  group_by(P.Product) %>%
  summarise(n_papers = n_distinct(B.Code), .groups = "drop") %>%
  arrange(desc(n_papers))

cat("\n✅ Summary by livestock product:\n")
print(ingredient_summary)


# ─────────────────────────────────────────────────────────────
# 3️⃣ Diet-based coverage
# ─────────────────────────────────────────────────────────────
# Use your current names: "diet_classif" (pairwise_diet_shift_tbl equivalent)
# and it contains column "Practice"
diet_classified_papers <- diet_classif %>%
  filter(!is.na(Subpractice)) %>%     # or another reliable diet label column
  distinct(B.Code)

diet_shift_papers <- diet_classif %>%
  filter(Practice == "Diet shift") %>%
  distinct(B.Code)

diet_with_yield <- GE_combined %>%
  filter(!is.na(CH4_kg_year)) %>%
  distinct(B.Code) %>%
  inner_join(diet_classified_papers, by = "B.Code") %>%
  inner_join(diet_shift_papers, by = "B.Code") %>%
  inner_join(yield_with_info, by = "B.Code") %>%
  distinct(B.Code)

diet_summary <- summarize_by_product(diet_with_yield, "diet_yield")

# ─────────────────────────────────────────────────────────────
# 4️⃣ Combine & Print
# ─────────────────────────────────────────────────────────────
total_ingredient <- n_distinct(ingredient_with_yield$B.Code)
total_diet <- n_distinct(diet_with_yield$B.Code)



cat("\n--- Ingredient-level coverage ---\n")
print(ingredient_summary)
cat("\n--- Diet-level coverage ---\n")
print(diet_summary)

```

## diag 

```{r}
library(dplyr)

# papers in the broad ingredient summary universe
papers_ingredient <- ingredient_with_yield %>% distinct(B.Code, P.Product)

# papers that actually made it into ΔEI (pair-level)
papers_delta <- final_with_deltas %>% distinct(B.Code, P.Product)

cat("Ingredient universe (papers):", nrow(papers_ingredient), "\n")
cat("ΔEI universe (papers):", nrow(papers_delta), "\n")

# which ingredient-eligible papers failed to produce ΔEI rows?
missing_in_delta <- anti_join(papers_ingredient, papers_delta, by = "B.Code")
cat("\nPapers present in ingredient summary but missing from ΔEI:\n")
print(missing_in_delta)

# quick reason flags (best-effort)
# 1) has at least one trt–ctrl pair in practice_tbl?
has_pairs <- practice_tbl %>% distinct(B.Code) %>% mutate(has_pairs = TRUE)

# 2) has CH4 per arm?
ch4_arm <- GE_combined %>%
  filter(!is.na(CH4_kg_year)) %>%
  transmute(B.Code, A.Level.Name = D.Item) %>%
  distinct() %>% mutate(has_ch4_arm = TRUE)

# 3) has yield per arm?
yield_arm <- yield_with_emissions %>%
  transmute(B.Code, A.Level.Name) %>%
  distinct() %>% mutate(has_yield_arm = TRUE)

# Cross-check at arm level
arm_diag <- full_join(ch4_arm, yield_arm, by = c("B.Code","A.Level.Name")) %>%
  mutate(has_both_arm = coalesce(has_ch4_arm, FALSE) & coalesce(has_yield_arm, FALSE)) %>%
  group_by(B.Code) %>%
  summarise(
    n_arms      = n(),
    n_arms_both = sum(has_both_arm, na.rm = TRUE),
    .groups = "drop"
  )

reasons <- missing_in_delta %>%
  left_join(has_pairs,  by = "B.Code") %>%
  left_join(arm_diag,   by = "B.Code") %>%
  mutate(
    has_pairs   = coalesce(has_pairs, FALSE),
    n_arms      = coalesce(n_arms, 0L),
    n_arms_both = coalesce(n_arms_both, 0L),
    likely_reason = dplyr::case_when(
      !has_pairs ~ "No trt–control pairs in practice_tbl",
      n_arms_both < 2 ~ "Missing aligned CH4+yield for at least two arms",
      TRUE ~ "Filtered later (units, EI calc, 'No change', or outlier)"
    )
  ) %>%
  arrange(likely_reason, P.Product, B.Code)

cat("\nLikely reasons for drops:\n")
print(reasons)

```

