---
title: "8_bis_Emission_Analysis_high_level"
output: html_document
date: "2025-10-28"
---

# A Productivity–Emission Relationship Analysis


To characterize the relationship between animal productivity and methane emissions across ruminant production systems, we compiled experimental data from controlled feeding studies reporting both daily product yield and annual methane output per animal. Each observation represented a discrete diet treatment within a study, identified by a unique study identifier code. Productivity metrics were standardized to edible product yield, expressed as either milk yield (kg animal⁻¹ day⁻¹) or meat yield derived from carcass composition or growth rate measurements. Methane emissions were standardized to annual production (kg CH₄ animal⁻¹ yr⁻¹).
Linear regression models were fitted independently for each species × product combination to quantify the relationship between productivity and absolute methane emissions. This approach captured within-study scaling relationships while maintaining comparability across experimental contexts. The analysis was restricted to genuine production outputs (milk or meat) to exclude potentially confounding metrics such as live weight gain or feed intake. Each regression coefficient therefore represents the empirical association between productivity and emission intensity across dietary treatments within comparable experimental frameworks.
The resulting relationships (Figure 1) illustrate how improvements in production output correspond to changes in total methane emission at the animal scale, providing cross-species visualization of productivity–emission trade-offs. By employing experimental contrasts rather than aggregated national inventory data, this analytical framework isolated biological variation in feed efficiency from management- or system-level effects.
```{r include=FALSE}

library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(ggplot2)
library(forcats)
library(scales)

GE_combined <- read.csv("~/Emission_Analysis/data/GE_Combined.csv")
merged_metadata <- readRDS("~/Emission_Analysis/data/merged_metadata_nutrition_inference.rds")
ingredients <- readRDS("~/Emission_Analysis/data/ingredients_harmonized_subset.rds")
yield_data<- read.csv("~/Emission_Analysis/data/yield_data.csv")%>%select(-c("P.Product"))
diet_percentages <- read.csv("~/Emission_Analysis/data/diet_percentages_classif.csv")
total_amounts <- read.csv("~/Emission_Analysis/data/total_amounts.csv")
ing_classif<-read.csv("~/Emission_Analysis/data/classif_ing_simple.csv")



```


```{r include=FALSE}
clean_key <- function(x) x %>% str_trim() %>% str_squish() %>% str_to_lower()
canonicalize_item <- function(x) {
  clean_key(x) %>%
    str_replace_all("\\bbran\\b", "offal")
}

  
  merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>%
  mutate(D.Item = canonicalize_item(D.Item)) 
  
    diet_percentages <-diet_percentages %>%
  mutate(D.Item = canonicalize_item(D.Item)) 

```



```{r echo=FALSE,fig.width=12, fig.height=7, out.width='100%', fig.align='center'}

# Step 1: Make sure the key columns match (D.Item and A.Level.Name)
GE_combined_for_merge <- GE_combined %>%
  filter(!is.na(CH4_kg_year))%>%
  rename(A.Level.Name = D.Item)%>%
  select(-"ED.Mean.T")


yield_data <- yield_data %>%
  filter(!is.na(ED.Mean.T))%>%
  filter(Out.Subind!="Meat Yield-Final Body Weight")


# Step 2: Merge yield and emission data
yield_with_emissions <- GE_combined_for_merge %>%
  left_join(
    yield_data,
    by = c("B.Code", "A.Level.Name")
  )%>%
    filter(!is.na(ED.Mean.T))%>%
  filter(!is.na(CH4_kg_year))%>%
  distinct()

  

# distinct B.Code per product
bcode_per_product <- yield_with_emissions %>%          # or yield_data
  distinct(P.Product, B.Code) %>%                      # keep one row per pair
  count(P.Product, name = "n_BCodes")                  # tally them

bcode_per_product


# Optional: ensure species have readable order
species_order <- c("Cattle", "Goat", "Sheep")  # adjust to match your data
yield_with_emissions <- yield_with_emissions %>%
  mutate(P.Product = factor(P.Product, levels = species_order))

yield_with_emissions <- yield_with_emissions %>%
  filter(
    # Keep only real milk or meat yield types
    (
      str_detect(Out.Subind, regex("Milk Yield", ignore_case = TRUE)) |
      str_detect(Out.Subind, regex("Meat Yield", ignore_case = TRUE))
    ) &
    !str_detect(Out.Subind, regex("Body Weight|Final", ignore_case = TRUE))
  ) %>%
  mutate(
    Product = case_when(
      str_detect(Out.Subind, regex("milk", ignore_case = TRUE)) ~ "Milk",
      str_detect(Out.Subind, regex("meat", ignore_case = TRUE)) ~ "Meat",
      TRUE ~ NA_character_
    )
  )


# Plot: one panel per species


ggplot(yield_with_emissions,
       aes(x = ED.Mean.T, y = CH4_kg_year)) +
  geom_point(
    color = "grey40", alpha = 0.35, size = 1.8
  ) +
  geom_smooth(
    method = "lm", se = TRUE,
    color = "grey20", linewidth = 0.5,
    fill = alpha("lightblue", 0.991)
  ) +
  facet_wrap(~ interaction(P.Product, Product, sep = " – "),
             scales = "free") +
  scale_y_continuous(
    "Annual CH₄ emissions (kg CH₄ animal⁻¹ yr⁻¹)",
    labels = comma_format()
  ) +
  scale_x_continuous(
    "Yield (kg product animal⁻¹ day⁻¹)",
    labels = comma_format()
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major = element_line(color = "grey92", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(face = "plain"),
    axis.text = element_text(color = "black"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "none"
  )

length(unique(yield_with_emissions$B.Code))


```
Figure 1 shows the relationship between annual enteric CH₄ emissions and product yield for cattle, goats, and sheep, separated by meat and milk production systems. Across all species, CH₄ emissions increase with yield, indicating that higher-producing animals generally emit more methane per year. However, the magnitude and slope of these relationships differ among species. Cattle have the highest absolute CH₄ emissions, reflecting their larger size and output levels, while goats and sheep emit less overall but follow similar positive trends. The slope of increase appears steeper in goats and sheep, suggesting that small changes in productivity are associated with relatively larger changes in CH₄ output in these smaller ruminants. These species-specific patterns highlight the diversity of emission–yield dynamics captured in African feeding trials and underscore the need to interpret CH₄ responses within the biological and management context of each species.

```{r include=FALSE}
# --- 2) Ensure we have diet_details and diet_shares (built from your data) ----
# If you already have these objects, this block will just reuse them.

# Keep only precise outcomes you compare within-paper
keep_outcomes <- c(
  "Milk Yield","Meat Yield","Meat Yield-Hot Carcass","Meat Yield-Cold Carcass",
  "Meat Yield-Slaughter Body","Meat Yield- Carcass","Meat Yield-Empty Carcass Meat Yield"
)

# Build yield_with_emissions_ranked if not present
if (!exists("yield_with_emissions_ranked")) {
  stopifnot(exists("yield_with_emissions"))  # you said this is available
  yield_with_emissions_ranked <- yield_with_emissions %>%
    filter(Out.Subind %in% keep_outcomes) %>%
    mutate(
      CH4_kg_day = CH4_kg_year/365,
      EI_kgCH4_per_kgProd = CH4_kg_day / ED.Mean.T
    ) %>%
    group_by(P.Product, Product, Out.Subind, B.Code) %>%
    mutate(
      yield_scaled = as.numeric(scale(ED.Mean.T)),
      ch4_scaled   = as.numeric(scale(CH4_kg_year))
    ) %>%
    ungroup()}

# Build diet_details from your ingredient shares
stopifnot(exists("diet_percentages"))
diet_details <- yield_with_emissions_ranked %>%
  select(P.Product, Product, Out.Subind, B.Code, A.Level.Name, ED.Mean.T,
         CH4_kg_year, CH4_kg_day, EI_kgCH4_per_kgProd, T.Control) %>%
  distinct() %>%
  left_join(
    diet_percentages %>%
      rename(
        Ingredient.Name = any_of(c("Ingredient.Name","D.Item")),
        Diet.Percent    = any_of(c("Diet.Percent","Percentage_of_Diet"))
      ) %>%
      mutate(Ingredient.Name = tolower(Ingredient.Name)) %>%
      select(B.Code, A.Level.Name, Ingredient.Name, Diet.Percent),
    by = c("B.Code","A.Level.Name")
  )

# Normalize to shares (sums to 1 per diet)
diet_shares <- diet_details %>%
  mutate(Ingredient.Name = tolower(Ingredient.Name)) %>%
  group_by(B.Code, Out.Subind, A.Level.Name) %>%
  mutate(share = Diet.Percent / sum(Diet.Percent, na.rm = TRUE)) %>%
  ungroup()

# --- 3) Read model-based nutrient data & standardize names (UPPERCASE) ----
# -> file: data/ingredients_high_level_model.csv
model_nutrients <- read_csv("data/ingredients_high_level_model.csv",
                            show_col_types = FALSE) %>%
  mutate(across(everything(), ~ ifelse(. == "", NA, .))) %>%
  mutate(across(where(is.character), tolower))

# normalize column names to lower, then rename to our standards
names(model_nutrients) <- tolower(names(model_nutrients))
model_nutrients <- model_nutrients %>%
  rename(D.Item = any_of(c("ingredient","d.item","item","feed","name"))) %>%
  mutate(D.Item = tolower(D.Item)) %>%
  # force nutrients to UPPERCASE names
  rename_with(~ toupper(.x), .cols = any_of(c("cp","ndf","adf","ee","ash","ge","me"))) %>%
  select(any_of(c("D.Item","CP","NDF","ADF","EE","ASH","GE","ME")))

# ============================================================
# 🧩 Empirical + Modeled Nutrient Composition (clean + compact)
# ============================================================

# --- 1️⃣ Extract empirical ingredient composition ----
ingredient_compo <- merged_metadata$Animals.Diet.Comp %>%
  filter(is_entire_diet == "FALSE") %>%
  mutate(D.Item = tolower(D.Item)) %>%
  filter(D.Item %in% tolower(diet_details$Ingredient.Name))




# --- 2️⃣ Combine empirical and modeled data ----
ingredient_compo_combined <- ingredient_compo %>%
  mutate(
    DC.Variable = toupper(DC.Variable),
    DC.Value = suppressWarnings(as.numeric(DC.Value))
  ) %>%
  filter(DC.Unit %in% c("g/kg", "MJ/kg"),
         DC.Variable %in% c("CP", "NDF", "ADF", "EE", "ASH", "GE", "ME")) %>%
  group_by(D.Item, DC.Variable) %>%
  summarise(DC.Value = mean(DC.Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = DC.Variable, values_from = DC.Value) %>%
  full_join(model_nutrients, by = "D.Item", suffix = c("", "_MODEL")) %>%
  mutate(
    CP  = coalesce(CP,  as.numeric(CP_MODEL)),
    NDF = coalesce(NDF, as.numeric(NDF_MODEL)),
    ADF = coalesce(ADF, as.numeric(ADF_MODEL)),
    EE  = coalesce(EE,  as.numeric(EE_MODEL)),
    ASH = coalesce(ASH, as.numeric(ASH_MODEL)),
    GE  = coalesce(GE,  as.numeric(GE_MODEL)),
    ME  = coalesce(ME,  as.numeric(ME_MODEL))
  ) %>%
  select(D.Item, CP, NDF, ADF, EE, ASH, GE, ME)

# --- 3️⃣ Add representative "molasses" nutrient values ----
molasses_means <- merged_metadata$Animals.Diet.Comp %>%
  filter(is_entire_diet == FALSE) %>%
  mutate(
    D.Item = tolower(D.Item),
    DC.Variable = toupper(DC.Variable),
    DC.Value = suppressWarnings(as.numeric(DC.Value))
  ) %>%
  filter(str_detect(D.Item, "^unspecified\\s*molasse")) %>%
  filter(
    DC.Unit %in% c("g/kg", "MJ/kg"),
    DC.Variable %in% c("CP", "NDF", "ADF", "EE", "ASH", "GE", "ME")
  ) %>%
  group_by(DC.Variable) %>%
  summarise(value = mean(DC.Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = DC.Variable, values_from = value) %>%
  mutate(D.Item = "molasses")

# Step 2️⃣ – Replace the GE value with the mean GE from *all* molasse-type ingredients
molasses_GE_mean <- merged_metadata$Animals.Diet.Comp %>%
  filter(is_entire_diet == FALSE) %>%
  mutate(D.Item = tolower(D.Item)) %>%
  filter(str_detect(D.Item, "molasse")) %>%
  mutate(DC.Variable = toupper(DC.Variable)) %>%
  filter(DC.Variable == "GE", DC.Unit %in% c("g/kg", "MJ/kg")) %>%
  summarise(GE_mean = mean(as.numeric(DC.Value), na.rm = TRUE)) %>%
  pull(GE_mean)

# Step 3️⃣ – Replace GE in the molasses table with the computed mean
molasses_means <- molasses_means %>%
  mutate(GE = molasses_GE_mean)


# Step 4️⃣ – Add or replace this entry in the full composition table
ingredient_compo_combined <- ingredient_compo_combined %>%
  filter(D.Item != "molasses") %>%
  bind_rows(molasses_means %>% select(names(ingredient_compo_combined)))


cat("✅ Combined empirical + modeled composition built, with 'molasses' added.\n")


cat("✅ 'molasses' nutrient values set to mean of all ingredients containing 'molasse'.\n")


# --- 6) Nutrient coverage across all (used) ingredients ----
nutrient_coverage <- ingredient_compo_combined %>%
  pivot_longer(cols = c(CP, NDF, ADF, EE, ASH, GE, ME),
               names_to = "DC.Variable", values_to = "DC.Value") %>%
  group_by(DC.Variable) %>%
  summarize(
    n_with_value = sum(!is.na(DC.Value)),
    n_total      = n(),
    coverage_pct = round(100 * n_with_value / n_total, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(coverage_pct))
cat("\n✅ Nutrient coverage (ingredients used in diets):\n")
print(nutrient_coverage)

# --- 7) Define non-nutritional ingredients to ignore for completeness checks ----
non_nutritional_ingredients <- c(
  "salt", "urea", "lime", "cement", "alkali", "sulphur",
  "mineral mix", "minerals and vitamins mix", "mineral lick",
  "dicalcium phosphate", "monocalcium phosphate",
  "bone ground", "protected fat", "supplement",
  "mixture", "commercial feed", "vitamins", "trace minerals",
  "naoh", "ca(oh)2", "phosphate", "ash", "limestone",
  "molasses mineral mix", "salt block"
)

# --- 8) Which ingredients have complete core nutrients? (CP,NDF,ADF,EE,ASH,GE) ----
ingredient_has_data <- ingredient_compo_combined %>%
  filter(!D.Item %in% non_nutritional_ingredients) %>%
  mutate(across(c(CP, NDF, ADF, EE, ASH, ME), ~ !is.na(.))) %>%
  rename_with(~ paste0("HAS_", .), c(CP, NDF, ADF, EE, ASH, ME)) %>%
  mutate(ALL_CORE = if_all(starts_with("HAS_"), ~ .x))

# --- 9) Build ingredient nutrient table for joining (prefix with nut_) ----
ingredient_nut <- ingredient_compo_combined %>%
  filter(!D.Item %in% non_nutritional_ingredients) %>%   # <-- drop non-nutritional items
  rename(
    nut_CP = CP, nut_NDF = NDF, nut_ADF = ADF, nut_EE = EE,
    nut_ASH = ASH, nut_GE = GE, nut_ME = ME
  )

# --- 10) Diet-level nutrient composition (weighted by ingredient share) ----
nut_cols_gkg <- c("nut_CP","nut_NDF","nut_ADF","nut_EE","nut_ASH")
nut_cols_mj  <- c("nut_ME","nut_GE")

diet_nutrient <- diet_shares %>%
  left_join(ingredient_nut, by = c("Ingredient.Name" = "D.Item")) %>%
  group_by(P.Product, Product, Out.Subind, B.Code, A.Level.Name,
           T.Control, ED.Mean.T, CH4_kg_day, EI_kgCH4_per_kgProd) %>%
  summarise(across(
      all_of(c(nut_cols_gkg, nut_cols_mj)),
      ~ if (all(is.na(.x))) NA_real_ else sum(share * .x, na.rm = TRUE),
      .names = "{.col}"),
    n_ing_with_nut = sum(
      !is.na(nut_CP) | !is.na(nut_NDF) | !is.na(nut_ME) |
      !is.na(nut_ADF) | !is.na(nut_EE) | !is.na(nut_ASH)),
    .groups = "drop")



cat("\n✅ Diet-level nutrient table built. N diets:", nrow(diet_nutrient), "\n")


# --- 11) Diet-level nutrient coverage with diagnostics ----

# 1️⃣ Compute diet coverage as before
diet_coverage <- diet_shares %>%
  mutate(Ingredient.Name = tolower(Ingredient.Name)) %>%
  filter(!Ingredient.Name %in% non_nutritional_ingredients) %>%
  left_join(
    ingredient_has_data %>%
      select(D.Item, ALL_CORE, HAS_CP, HAS_NDF, HAS_ADF, HAS_EE, HAS_ASH, HAS_ME),
    by = c("Ingredient.Name" = "D.Item")
  ) %>%
  group_by(B.Code, Out.Subind, P.Product, Product, A.Level.Name) %>%
  summarize(
    n_ing_total = n(),
    n_ing_with_all_nut = sum(ALL_CORE, na.rm = TRUE),
    pct_with_data = round(100 * n_ing_with_all_nut / n_ing_total, 1),
    .groups = "drop"
  )

cat("\n📊 Diet coverage computed for", nrow(diet_coverage), "diets.\n")

# 2️⃣ Identify incomplete diets (<100% nutrient coverage)
incomplete_diets <- diet_coverage %>%
  filter(pct_with_data < 100)

cat("\n❗ Incomplete diets found:", nrow(incomplete_diets), "\n")

# 3️⃣ Diagnose *which ingredients* are missing which nutrients
diet_missing_details <- diet_shares %>%
  mutate(Ingredient.Name = tolower(Ingredient.Name)) %>%
  filter(!Ingredient.Name %in% non_nutritional_ingredients) %>%
  left_join(
    ingredient_compo_combined %>%
      select(D.Item, CP, NDF, ADF, EE, ASH, ME),
    by = c("Ingredient.Name" = "D.Item")
  ) %>%
  mutate(
    missing_CP  = is.na(CP),
    missing_NDF = is.na(NDF),
    missing_ADF = is.na(ADF),
    missing_EE  = is.na(EE),
    missing_ASH = is.na(ASH),
    missing_ME  = is.na(ME),
    missing_any = missing_CP | missing_NDF | missing_ADF | missing_EE | missing_ASH | missing_ME
  ) %>%
  filter(missing_any) %>%
  select(
    B.Code, Out.Subind, Product, A.Level.Name, Ingredient.Name,
    missing_CP, missing_NDF, missing_ADF, missing_EE, missing_ASH, missing_ME
  ) %>%
  arrange(B.Code, A.Level.Name, Ingredient.Name)

# 4️⃣ Summary of missing ingredients per diet
missing_summary <- diet_missing_details %>%
  group_by(B.Code, A.Level.Name) %>%
  summarize(
    n_missing_ingredients = n_distinct(Ingredient.Name),
    .groups = "drop"
  ) %>%
  left_join(
    diet_coverage %>% select(B.Code, A.Level.Name, pct_with_data),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  arrange(B.Code, desc(pct_with_data))

# 5️⃣ Display results
cat("\n🔍 Missing nutrient details (first 15 rows):\n")
print(head(diet_missing_details, 15))

cat("\n📈 Summary of incomplete diets:\n")
print(missing_summary)

# 6️⃣ Fully covered diets (100% of relevant ingredients have all nutrients)
complete_diets <- diet_coverage %>% filter(pct_with_data == 100)
cat("\n✅ Fully covered diets (relevant ingredients only):", nrow(complete_diets), "\n")

# Optional: Export diagnostics
# write_csv(diet_missing_details, "outputs/diet_missing_details.csv")
# write_csv(missing_summary, "outputs/diet_missing_summary.csv")



complete_diets <- diet_coverage %>% filter(pct_with_data == 100)
cat("\n✅ Fully covered diets (relevant ingredients only):", nrow(complete_diets), "\n")


# --- 12) Reference diet per paper × outcome (control if present, else lowest EI) ----
ref_diets <- diet_nutrient %>%
  group_by(B.Code, Out.Subind) %>%
  mutate(any_control = any(isTRUE(T.Control))) %>%
  ungroup() %>%
  group_by(B.Code, Out.Subind, A.Level.Name) %>%
  summarize(
    EI = first(EI_kgCH4_per_kgProd),
    CH4 = first(CH4_kg_day),
    Y   = first(ED.Mean.T),
    is_control = any(isTRUE(T.Control)),
    .groups = "drop"
  ) %>%
  group_by(B.Code, Out.Subind) %>%
  mutate(
    EI_rank = rank(EI, ties.method = "first"),
    ref_flag = if (any(is_control)) is_control else EI_rank == 1L
  ) %>%
  filter(ref_flag) %>%
  arrange(B.Code, Out.Subind, A.Level.Name) %>%
  slice_head(n = 1) %>%                    # ensure 1 reference per study×outcome
  ungroup()

# Bring reference nutrient composition and rename to *_ref
ref_full <- ref_diets %>%
  left_join(
    diet_nutrient %>%
      select(B.Code, Out.Subind, A.Level.Name, all_of(c(nut_cols_gkg, nut_cols_mj))),
    by = c("B.Code","Out.Subind","A.Level.Name")
  ) %>%
  rename(A.Level.Name_ref = A.Level.Name) %>%
  rename(
    EI_ref  = EI,
    CH4_ref = CH4,
    Y_ref   = Y
  ) %>%
  rename_with(~ paste0(.x, "_ref"), all_of(c(nut_cols_gkg, nut_cols_mj)))

# --- 13) Within-study deltas vs reference (no many-to-many join issues) ----
diet_deltas <- diet_nutrient %>%
  left_join(ref_full, by = c("B.Code","Out.Subind")) %>%
  filter(A.Level.Name != A.Level.Name_ref) %>%   # drop the reference rows
  mutate(
    dY    = ED.Mean.T - Y_ref,
    dCH4  = CH4_kg_day - CH4_ref,
    dEI   = EI_kgCH4_per_kgProd - EI_ref,
    dY_pct   = 100 * dY / Y_ref,
    dCH4_pct = 100 * dCH4 / CH4_ref,
    dEI_pct  = 100 * dEI / EI_ref
  ) %>%
  # nutrient deltas (absolute)
  mutate(
    across(all_of(c(nut_cols_gkg, nut_cols_mj)),
           ~ . - get(paste0(cur_column(), "_ref")), .names = "d{.col}")
  )

cat("\n✅ Delta table built. N within-study pairs:", nrow(diet_deltas), "\n")

# --- 14) Quick reportable counts ----
counts <- list(
  diets_total      = nrow(diet_nutrient),
  papers_total     = dplyr::n_distinct(diet_nutrient$B.Code),
  outcomes_total   = dplyr::n_distinct(diet_nutrient$Out.Subind),
  pairs_total      = nrow(diet_deltas),
  pairs_with_ME    = sum(!is.na(diet_deltas$dnut_ME)),
  pairs_with_NDF   = sum(!is.na(diet_deltas$dnut_NDF)),
  pairs_with_CP    = sum(!is.na(diet_deltas$dnut_CP)),
  pairs_with_EE    = sum(!is.na(diet_deltas$dnut_EE))
)
cat("\n✅ Counts:\n"); print(counts)

# --- 15) FIGURE 1 – Coefficient plot (compact) ----
# Uses base lm; broom::tidy() via namespace to avoid attach
# --- 15) MODELS + COUNTS (EI, Yield, CH4) -------------------------------

# helper: rescale nutrient deltas to reporting units
scale_vars <- function(df) {
  df %>%
    mutate(
      dnut_ME_1    = dnut_ME,        # per 1 MJ/kg
      dnut_NDF_100 = dnut_NDF / 100, # per 100 g/kg
      dnut_CP_100  = dnut_CP  / 50, # per 50g/kg
      dnut_EE_10   = dnut_EE  / 10   # per 10 g/kg
    )
}

# build a base modeling table; DON'T require all three outcomes simultaneously
base_data <- diet_deltas %>%
  scale_vars() %>%
  mutate(.row_id = dplyr::row_number()) %>%
  # keep ids you'll use for study counts
  select(.row_id, B.Code, P.Product, Product,
         dEI_pct, dY_pct, dCH4_pct,
         dnut_ME_1, dnut_NDF_100, dnut_CP_100, dnut_EE_10)

# per-outcome modeling data (separate filters)
mod_ei_data  <- base_data %>% filter(!is.na(dEI_pct))
mod_y_data   <- base_data %>% filter(!is.na(dY_pct))
mod_ch4_data <- base_data %>% filter(!is.na(dCH4_pct))

# masks of rows actually used (complete cases on predictors + response)
preds <- c("dnut_ME_1","dnut_NDF_100","dnut_CP_100","dnut_EE_10","P.Product","Product")

mask_ei  <- stats::complete.cases(mod_ei_data[, c("dEI_pct",  preds)])
mask_y   <- stats::complete.cases(mod_y_data[,  c("dY_pct",   preds)])
mask_ch4 <- stats::complete.cases(mod_ch4_data[,c("dCH4_pct", preds)])

# fit models
m_ei  <- lm(dEI_pct  ~ dnut_ME_1 + dnut_NDF_100 + dnut_CP_100 + dnut_EE_10 + P.Product + Product,
            data = mod_ei_data[mask_ei, ])
m_y   <- lm(dY_pct   ~ dnut_ME_1 + dnut_NDF_100 + dnut_CP_100 + dnut_EE_10 + P.Product + Product,
            data = mod_y_data[mask_y, ])
m_ch4 <- lm(dCH4_pct ~ dnut_ME_1 + dnut_NDF_100 + dnut_CP_100 + dnut_EE_10 + P.Product + Product,
            data = mod_ch4_data[mask_ch4, ])

# counts: datapoints (rows) and papers (distinct B.Code) per model
model_meta <- dplyr::tibble(
  outcome   = c("ΔEI (%)","ΔYield (%)","ΔCH₄ (%)"),
  n_obs     = c(sum(mask_ei), sum(mask_y), sum(mask_ch4)),
  n_papers  = c(
    dplyr::n_distinct(mod_ei_data$B.Code[mask_ei]),
    dplyr::n_distinct(mod_y_data$B.Code[mask_y]),
    dplyr::n_distinct(mod_ch4_data$B.Code[mask_ch4])
  )
)

cat("Obs used — EI:", model_meta$n_obs[1],
    "| Yield:", model_meta$n_obs[2],
    "| CH4:", model_meta$n_obs[3], "\n")
cat("Papers used — EI:", model_meta$n_papers[1],
    "| Yield:", model_meta$n_papers[2],
    "| CH4:", model_meta$n_papers[3], "\n")

# --- 16) COEFFICIENT TABLE FOR PLOT -------------------------------------

# --- 16) COEFFICIENT TABLE FOR PLOT (with n_obs and n_papers per nutrient) ----

coef_df <- function(m, label, data, outcome) {
  # keep only the actual rows used for the model
  complete_data <- data[stats::complete.cases(data[, c(outcome,
                                                      "dnut_ME_1","dnut_NDF_100",
                                                      "dnut_CP_100","dnut_EE_10",
                                                      "P.Product","Product")]), ]

  broom::tidy(m, conf.int = TRUE) %>%
    dplyr::filter(term %in% c("dnut_ME_1","dnut_NDF_100","dnut_CP_100","dnut_EE_10")) %>%
    dplyr::mutate(
      outcome = label,
      term = dplyr::recode(term,
        "dnut_ME_1"    = "ΔME (per 1 MJ/kg)",
        "dnut_NDF_100" = "ΔNDF (per 100 g/kg)",
        "dnut_CP_100"  = "ΔCP (per 50 g/kg)",
        "dnut_EE_10"   = "ΔEE (per 10 g/kg)"
      ),
      n_obs = nrow(complete_data),
      n_papers = dplyr::n_distinct(complete_data$B.Code)
    )
}

coef_all <- dplyr::bind_rows(
  coef_df(m_ei,  "ΔEI (%)",   mod_ei_data,  "dEI_pct"),
  coef_df(m_y,   "ΔYield (%)", mod_y_data,  "dY_pct"),
  coef_df(m_ch4, "ΔCH4 (%)",  mod_ch4_data, "dCH4_pct")
)

# --- 17) Pretty coefficient table ----
coef_table <- coef_all %>%
  dplyr::select(outcome, term, estimate, conf.low, conf.high,
                std.error, statistic, p.value, n_obs, n_papers) %>%
  dplyr::arrange(outcome, term)

cat("\n📊 Coefficient table (with sample size and papers per term):\n")
print(coef_table, n = Inf)



# --- 17) FACET LABELS WITH COUNTS ---------------------------------------

# build facet labels "Outcome (n=..., papers=...)"
lab_map <- setNames(
  paste0(model_meta$outcome, " (n=", model_meta$n_obs, ", papers=", model_meta$n_papers, ")"),
  model_meta$outcome
)

# --- 18) PLOT ------------------------------------------------------------
```


```{r include=FALSE}
ggplot(coef_all,
       aes(x = estimate, y = forcats::fct_rev(term), xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_errorbarh(height = 0) +
  geom_point(size = 2) +
  facet_wrap(~ outcome, scales = "free_x",
             labeller = labeller(outcome = lab_map)) +
  labs(
    title = "Within-study nutrient shifts vs. ΔEI, ΔYield, and ΔCH4",
    subtitle = "Facet labels show datapoints (n) and distinct papers used",
    x = "Estimated effect (95% CI)",
    y = NULL,
    caption = paste0("EI n=", model_meta$n_obs[1], " (", model_meta$n_papers[1], " papers); ",
                     "Yield n=", model_meta$n_obs[2], " (", model_meta$n_papers[2], " papers); ",
                     "CH4 n=", model_meta$n_obs[3], " (", model_meta$n_papers[3], " papers).")
  ) +
  theme_minimal(base_size = 13)



```


```{r include=FALSE}
# ===== A) LONG TABLE (one row per term × outcome) =====
# Assumes `coef_all` already exists from your plotting code.

coef_table_long <- coef_all %>%
  dplyr::select(outcome, term, estimate, conf.low, conf.high, std.error, p.value) %>%
  dplyr::mutate(
    dplyr::across(c(estimate, conf.low, conf.high, std.error), ~ round(., 3)),
    p.value = signif(p.value, 3)
  ) %>%
  dplyr::rename(
    Outcome = outcome,
    `Nutrient term` = term,
    Estimate = estimate,
    `CI low` = conf.low,
    `CI high` = conf.high,
    `Std. Error` = std.error,
    `p-value` = p.value
  ) %>%
  dplyr::arrange(Outcome, `Nutrient term`)

# Print nicely (optional)
if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::kable(coef_table_long, align = "l", caption = "Coefficients used in the plot")
} else {
  print(coef_table_long, n = Inf)
}


```

```{r include=FALSE}
# --- Mean within-paper nutrient changes + distributions -----------------------


# 1️⃣ Remove zero or missing nutrient deltas first
diet_deltas_nonzero <- diet_deltas %>%
  mutate(across(starts_with("dnut_"),
                ~ ifelse(.x == 0, NA, .x))) %>%  # turn zeros to NA
  filter(
    !(is.na(dnut_CP) & is.na(dnut_NDF) & is.na(dnut_ADF) &
      is.na(dnut_EE) & is.na(dnut_ME))
  )

# 2️⃣ Compute mean absolute nutrient changes per paper × outcome
nut_delta_summary <- diet_deltas_nonzero %>%
  group_by(B.Code, Out.Subind) %>%
  summarise(
    across(
      c(dnut_CP, dnut_NDF, dnut_ADF, dnut_EE, dnut_ME),
      ~ mean(abs(.x), na.rm = TRUE),
      .names = "mean_abs_{.col}"
    ),
    .groups = "drop"
  ) %>%
  # 3️⃣ Summarise across all papers to get global averages and spread
  summarise(
    across(
      starts_with("mean_abs_"),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        sd = ~ sd(.x, na.rm = TRUE),
        median = ~ median(.x, na.rm = TRUE),
        q25 = ~ quantile(.x, 0.25, na.rm = TRUE),
        q75 = ~ quantile(.x, 0.75, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )

# 4️⃣ Print compact table with units
cat("\n📊 Mean within-paper absolute nutrient changes (excluding zeros):\n")
nut_table <- data.frame(
  Nutrient = c("CP (g/kg DM)", "NDF (g/kg DM)", "ADF (g/kg DM)", "EE (g/kg DM)", "ME (MJ/kg DM)"),
  Mean = round(c(nut_delta_summary$mean_abs_dnut_CP_mean,
                 nut_delta_summary$mean_abs_dnut_NDF_mean,
                 nut_delta_summary$mean_abs_dnut_ADF_mean,
                 nut_delta_summary$mean_abs_dnut_EE_mean,
                 nut_delta_summary$mean_abs_dnut_ME_mean), 2),
  SD = round(c(nut_delta_summary$mean_abs_dnut_CP_sd,
               nut_delta_summary$mean_abs_dnut_NDF_sd,
               nut_delta_summary$mean_abs_dnut_ADF_sd,
               nut_delta_summary$mean_abs_dnut_EE_sd,
               nut_delta_summary$mean_abs_dnut_ME_sd), 2),
  Median = round(c(nut_delta_summary$mean_abs_dnut_CP_median,
                   nut_delta_summary$mean_abs_dnut_NDF_median,
                   nut_delta_summary$mean_abs_dnut_ADF_median,
                   nut_delta_summary$mean_abs_dnut_EE_median,
                   nut_delta_summary$mean_abs_dnut_ME_median), 2),
  Q25 = round(c(nut_delta_summary$mean_abs_dnut_CP_q25,
                nut_delta_summary$mean_abs_dnut_NDF_q25,
                nut_delta_summary$mean_abs_dnut_ADF_q25,
                nut_delta_summary$mean_abs_dnut_EE_q25,
                nut_delta_summary$mean_abs_dnut_ME_q25), 2),
  Q75 = round(c(nut_delta_summary$mean_abs_dnut_CP_q75,
                nut_delta_summary$mean_abs_dnut_NDF_q75,
                nut_delta_summary$mean_abs_dnut_ADF_q75,
                nut_delta_summary$mean_abs_dnut_EE_q75,
                nut_delta_summary$mean_abs_dnut_ME_q75), 2)
)
print(nut_table)

# 5️⃣ Histogram plot (excluding zeros)
diet_deltas_nonzero %>%
  pivot_longer(c(dnut_CP, dnut_NDF, dnut_EE, dnut_ME, dnut_ADF),
               names_to = "nutrient", values_to = "delta") %>%
  filter(!is.na(delta) & delta != 0) %>%
  mutate(nutrient = str_remove(nutrient, "dnut_")) %>%
  ggplot(aes(x = abs(delta))) +
  geom_histogram(bins = 30, fill = "grey70", color = "white") +
  facet_wrap(~ nutrient, scales = "free_x") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Distribution of within-study nutrient differences across diets",
    subtitle = "Zero nutrient changes removed",
    x = "Absolute within-study nutrient change",
    y = "Count"
  )

```
```{r include=FALSE}
# --- Mean and absolute within-study nutrient changes --------------------------
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)

# 1️⃣ Remove zero or missing nutrient deltas (keep only diets that changed)
diet_deltas_nonzero <- diet_deltas %>%
  mutate(across(starts_with("dnut_"), ~ ifelse(.x == 0, NA, .x))) %>%
  filter(
    !(is.na(dnut_CP) & is.na(dnut_NDF) & is.na(dnut_ADF) &
      is.na(dnut_EE) & is.na(dnut_ME))
  )

# 2️⃣ Compute within-study nutrient changes (mean per paper × outcome)
nut_deltas_summary <- diet_deltas_nonzero %>%
  group_by(B.Code, Out.Subind) %>%
  summarise(
    across(
      c(dnut_CP, dnut_NDF, dnut_ADF, dnut_EE, dnut_ME),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        abs_mean = ~ mean(abs(.x), na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )

# 3️⃣ Aggregate across studies to global summaries (mean, sd, median, IQR)
nut_delta_global <- nut_deltas_summary %>%
  select(where(is.numeric)) %>%
  summarise(
    across(
      everything(),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        sd   = ~ sd(.x, na.rm = TRUE),
        median = ~ median(.x, na.rm = TRUE),
        q25 = ~ quantile(.x, 0.25, na.rm = TRUE),
        q75 = ~ quantile(.x, 0.75, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )

# 4️⃣ Build compact summary tables
## Absolute changes
nut_table_abs <- data.frame(
  Nutrient = c("CP (g/kg DM)", "NDF (g/kg DM)", "ADF (g/kg DM)",
               "EE (g/kg DM)", "ME (MJ/kg DM)"),
  Mean_abs = round(c(nut_delta_global$dnut_CP_abs_mean_mean,
                     nut_delta_global$dnut_NDF_abs_mean_mean,
                     nut_delta_global$dnut_ADF_abs_mean_mean,
                     nut_delta_global$dnut_EE_abs_mean_mean,
                     nut_delta_global$dnut_ME_abs_mean_mean), 2),
  SD_abs = round(c(nut_delta_global$dnut_CP_abs_mean_sd,
                   nut_delta_global$dnut_NDF_abs_mean_sd,
                   nut_delta_global$dnut_ADF_abs_mean_sd,
                   nut_delta_global$dnut_EE_abs_mean_sd,
                   nut_delta_global$dnut_ME_abs_mean_sd), 2),
  Median_abs = round(c(nut_delta_global$dnut_CP_abs_mean_median,
                       nut_delta_global$dnut_NDF_abs_mean_median,
                       nut_delta_global$dnut_ADF_abs_mean_median,
                       nut_delta_global$dnut_EE_abs_mean_median,
                       nut_delta_global$dnut_ME_abs_mean_median), 2),
  Q25_abs = round(c(nut_delta_global$dnut_CP_abs_mean_q25,
                    nut_delta_global$dnut_NDF_abs_mean_q25,
                    nut_delta_global$dnut_ADF_abs_mean_q25,
                    nut_delta_global$dnut_EE_abs_mean_q25,
                    nut_delta_global$dnut_ME_abs_mean_q25), 2),
  Q75_abs = round(c(nut_delta_global$dnut_CP_abs_mean_q75,
                    nut_delta_global$dnut_NDF_abs_mean_q75,
                    nut_delta_global$dnut_ADF_abs_mean_q75,
                    nut_delta_global$dnut_EE_abs_mean_q75,
                    nut_delta_global$dnut_ME_abs_mean_q75), 2)
)

## Signed changes
nut_table_signed <- data.frame(
  Nutrient = c("CP (g/kg DM)", "NDF (g/kg DM)", "ADF (g/kg DM)",
               "EE (g/kg DM)", "ME (MJ/kg DM)"),
  Mean_signed = round(c(nut_delta_global$dnut_CP_mean_mean,
                        nut_delta_global$dnut_NDF_mean_mean,
                        nut_delta_global$dnut_ADF_mean_mean,
                        nut_delta_global$dnut_EE_mean_mean,
                        nut_delta_global$dnut_ME_mean_mean), 2),
  SD_signed = round(c(nut_delta_global$dnut_CP_mean_sd,
                      nut_delta_global$dnut_NDF_mean_sd,
                      nut_delta_global$dnut_ADF_mean_sd,
                      nut_delta_global$dnut_EE_mean_sd,
                      nut_delta_global$dnut_ME_mean_sd), 2),
  Median_signed = round(c(nut_delta_global$dnut_CP_mean_median,
                          nut_delta_global$dnut_NDF_mean_median,
                          nut_delta_global$dnut_ADF_mean_median,
                          nut_delta_global$dnut_EE_mean_median,
                          nut_delta_global$dnut_ME_mean_median), 2),
  Q25_signed = round(c(nut_delta_global$dnut_CP_mean_q25,
                       nut_delta_global$dnut_NDF_mean_q25,
                       nut_delta_global$dnut_ADF_mean_q25,
                       nut_delta_global$dnut_EE_mean_q25,
                       nut_delta_global$dnut_ME_mean_q25), 2),
  Q75_signed = round(c(nut_delta_global$dnut_CP_mean_q75,
                       nut_delta_global$dnut_NDF_mean_q75,
                       nut_delta_global$dnut_ADF_mean_q75,
                       nut_delta_global$dnut_EE_mean_q75,
                       nut_delta_global$dnut_ME_mean_q75), 2)
)

# 5️⃣ Print both tables
cat("\n📊 Mean absolute within-study nutrient changes (excluding zeros):\n")
print(nut_table_abs)
cat("\n📈 Mean signed within-study nutrient changes (directional, excluding zeros):\n")
print(nut_table_signed)

# 6️⃣ Optional histogram for visual check (zeros removed)
diet_deltas_nonzero %>%
  pivot_longer(c(dnut_CP, dnut_NDF, dnut_EE, dnut_ME, dnut_ADF),
               names_to = "nutrient", values_to = "delta") %>%
  filter(!is.na(delta) & delta != 0) %>%
  mutate(nutrient = str_remove(nutrient, "dnut_")) %>%
  ggplot(aes(x = delta)) +
  geom_histogram(bins = 30, fill = "grey75", color = "white") +
  facet_wrap(~ nutrient, scales = "free_x") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Distribution of within-study nutrient shifts (zeros excluded)",
    x = "Within-study change (signed, g/kg or MJ/kg DM)",
    y = "Count"
  )

```

# B Nutrient–Response Model Development

## Within-Study Comparative Framework


To quantify the effects of dietary composition on productivity and methane emissions, we employed a within-study comparative framework that controlled for study-specific experimental conditions. For each study included in the emission–yield database, the diet designated as the reference or control treatment was used as the baseline. All alternative dietary treatments within that study were expressed relative to this reference, yielding within-study percentage changes in methane output (ΔCH₄ %), product yield (ΔYield %), and calculated energy intake (ΔEI %).
Corresponding within-study changes in dietary composition were computed for metabolizable energy (ΔME), neutral detergent fiber (ΔNDF), crude protein (ΔCP), and ether extract (ΔEE). Nutrient concentrations were standardized to comparable reporting units: MJ kg⁻¹ dry matter (DM) for ΔME and g kg⁻¹ DM for ΔNDF, ΔCP, and ΔEE. This normalization procedure ensured that all regression coefficients represented biological responses to nutritional variation observed within identical experimental contexts, thereby minimizing confounding effects associated with between-study heterogeneity in animal genetics, environmental conditions, and management practices.

##Statistical Modeling Approach

Separate linear regression models were fitted for each combination of species (cattle, sheep, goat) and production system (meat or milk). Within each subset, changes in methane emissions, energy intake, and product yield were modeled as linear functions of the nutrient deltas, with independent slope coefficients estimated for each nutrient–outcome pair. These coefficients quantify the proportional change in response variables (methane, intake, or yield) associated with unit increases in dietary nutrient concentrations relative to the reference diet. Only species × production system combinations with sufficient within-study dietary contrasts were retained to ensure reliable parameter estimation.

## Presentation of Species-Specific Response Patterns

Figures 3 and 4 present estimated coefficients as species-specific response patterns for meat and milk production systems, respectively. Each column corresponds to a species and each row to an outcome variable (ΔCH₄ %, ΔYield %, or ΔEI %). Independent x-axis scaling was applied across panels to preserve the true magnitude of effects and enable comparison among systems.

Nutrient predictors were standardized to realistic unit changes—1 MJ/kg for metabolizable energy (ME), 100 g/kg for neutral detergent fibre (NDF), 50 g/kg for crude protein (CP), and 10 g/kg for ether extract (EE)—based on the median within-study diet contrasts observed in our dataset. These increments reflect typical formulation ranges used in African feeding trials and ensure that coefficients correspond to nutritionally meaningful differences.
```{r,echo=FALSE,fig.width=12, fig.height=7, out.width='100%', fig.align='center'}
# ============================================================
# 🔍 Within-study nutrient shifts vs ΔEI, ΔYield, and ΔCH₄
#     One column per Species (Cattle, Sheep, Goat)
#     Nutrient variables shown once (left), Δ outcomes shown once (right)
# ============================================================

library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(broom)
library(purrr)
library(patchwork)
library(ggforce)  # for flexible faceting

# 0) Safety checks ---------------------------------------------------------
stopifnot(exists("diet_deltas"))

# 1) Scale helper ----------------------------------------------------------
scale_vars <- function(df) {
  df %>%
    mutate(
      dnut_ME_1    = dnut_ME,         # per 1 MJ/kg
      dnut_NDF_100 = dnut_NDF / 100,  # per 100 g/kg
      dnut_CP_50   = dnut_CP  / 50,   # per 50 g/kg
      dnut_EE_10   = dnut_EE  / 10    # per 10 g/kg
    )
}

# 2) Base data -------------------------------------------------------------
base_data <- diet_deltas %>%
  scale_vars() %>%
  transmute(
    B.Code,
    Species = P.Product,
    Product,
    dEI_pct, dY_pct, dCH4_pct,
    dnut_ME_1, dnut_NDF_100, dnut_CP_50, dnut_EE_10
  ) %>%
  mutate(
    Species = factor(Species, levels = c("Cattle", "Sheep", "Goat")),
    Product = factor(Product, levels = c("Milk", "Meat"))
  )

# 3) Fit models ------------------------------------------------------------
fit_one <- function(sp, prd, n_min = 20) {
  dat <- base_data %>% filter(Species == sp, Product == prd)
  if (nrow(dat) < n_min) return(NULL)

  fit_outcome <- function(y, label) {
    preds <- c("dnut_ME_1","dnut_NDF_100","dnut_CP_50","dnut_EE_10")
    keep  <- stats::complete.cases(dat[, c(y, preds)])
    d     <- dat[keep, , drop = FALSE]
    if (nrow(d) < n_min) return(NULL)

    m <- lm(stats::as.formula(
      paste0(y, " ~ dnut_ME_1 + dnut_NDF_100 + dnut_CP_50 + dnut_EE_10")), data = d)

    broom::tidy(m, conf.int = TRUE) %>%
      filter(term %in% preds) %>%
      mutate(
        Species = sp,
        Product = prd,
        outcome = label,
        term = recode(term,
                      "dnut_ME_1"    = "ΔME (per 1 MJ/kg)",
                      "dnut_NDF_100" = "ΔNDF (per 100 g/kg)",
                      "dnut_CP_50"   = "ΔCP (per 50 g/kg)",
                      "dnut_EE_10"   = "ΔEE (per 10 g/kg)")
      )
  }

  bind_rows(
    fit_outcome("dEI_pct",  "ΔEI (%)"),
    fit_outcome("dY_pct",   "ΔYield (%)"),
    fit_outcome("dCH4_pct", "ΔCH4 (%)")
  )
}

# 4) Run across all Species × Product combos -------------------------------
all_combos <- tidyr::expand_grid(
  Species = levels(base_data$Species),
  Product = levels(base_data$Product)
)

coef_species_prod <- purrr::pmap_dfr(
  list(all_combos$Species, all_combos$Product),
  ~ fit_one(..1, ..2)
)

if (nrow(coef_species_prod) == 0) {
  stop("No Species × Product combinations had enough complete data to fit models.")
}

# 5) Prepare data ----------------------------------------------------------
# ============================================================
# 🔍 Within-study nutrient shifts vs. ΔEI, ΔYield, and ΔCH₄
#     One column per Species (Cattle, Sheep, Goat)
#     Nutrient variables shown once on left, Δ labels once on right
# ============================================================

library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(broom)
library(purrr)
library(patchwork)

# --- assume coef_species_prod already created as before ---

# Ensure factors are tidy and consistent
coef_species_prod <- coef_species_prod %>%
  mutate(
    outcome = factor(outcome, levels = c("ΔCH4 (%)", "ΔEI (%)", "ΔYield (%)")),
    panel   = interaction(Product, outcome, sep = " – ", drop = TRUE)
  )

# Split by species
cattle_df <- coef_species_prod %>% filter(Species == "Cattle")
sheep_df  <- coef_species_prod %>% filter(Species == "Sheep")
goat_df   <- coef_species_prod %>% filter(Species == "Goat")

# ------------------------------------------------------------------
# Plot function: free x-scales per small panel, hide duplicate labels
# ------------------------------------------------------------------
# Define custom soft, semi-transparent colors
nutrient_colors <- c(
  "ΔCP (per 50 g/kg)"  = "blue4",
  "ΔEE (per 10 g/kg)"  = "darkorange",
  "ΔME (per 1 MJ/kg)"  = "seagreen3",
  "ΔNDF (per 100 g/kg)"= "plum3"
)

# Updated plotting function
plot_species <- function(data, sp_name, show_y = TRUE, show_right = FALSE) {
  ggplot(data,
         aes(x = estimate, y = fct_rev(term), color = term)) +
    geom_vline(xintercept = 0, linetype = 2, color = "grey70") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0.15, linewidth = 1.1, alpha = 0.7) +
    geom_point(size = 3.2, alpha = 0.8) +
    facet_grid(
      rows = vars(outcome),
      cols = vars(Product),
      scales = "free_x",
      switch = if (show_right) "both" else "y"
    ) +
    scale_color_manual(values = nutrient_colors, name = "Nutrient variable") +
    labs(
      subtitle = sp_name,
      x = "Estimated effect (95% CI)",
      y = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      strip.text.x  = element_text(face = "bold", size = 11),
      strip.text.y  = element_text(face = "bold", size = 11),
      strip.placement = "outside",
      panel.spacing = unit(1.2, "lines"),
      plot.subtitle = element_text(face = "bold", size = 13, hjust = 0.5),
      axis.text.y   = if (show_y) element_text(face = "bold") else element_blank(),
      axis.ticks.y  = if (show_y) element_line() else element_blank(),
      legend.position = "none"
    )
}


# ------------------------------------------------------------------
# Build species plots
# ------------------------------------------------------------------
p_cattle <- plot_species(cattle_df, "Cattle", show_y = TRUE,  show_right = FALSE)
p_sheep  <- plot_species(sheep_df,  "Sheep",  show_y = FALSE, show_right = FALSE)
p_goat   <- plot_species(goat_df,   "Goat",   show_y = FALSE, show_right = TRUE)

# ------------------------------------------------------------------
# Combine horizontally with patchwork
# ------------------------------------------------------------------
final_plot <- p_cattle | p_sheep | p_goat +
  plot_annotation(
    title = "Within-study nutrient shifts vs. ΔEI, ΔYield, and ΔCH₄",
  )

final_plot


```

```{r,echo=FALSE,fig.width=12, fig.height=7, out.width='100%', fig.align='center'}
# ============================================================
# 🔍 Within-study nutrient shifts vs ΔEI, ΔYield, and ΔCH₄ (Milk only)
#     One column per Species (Cattle, Sheep, Goat)
#     Nutrient variables shown once (left), Δ outcomes once (right)
# ============================================================

library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(broom)
library(purrr)
library(patchwork)

# 1️⃣ Prepare scaled data --------------------------------------------------
base_data_milk <- diet_deltas %>%
  mutate(
    Species = factor(P.Product, levels = c("Cattle", "Sheep", "Goat")),
    Product = factor(Product, levels = c("Milk", "Meat"))
  ) %>%
  filter(Product == "Milk") %>%   # keep Milk systems only
  mutate(
    dnut_ME_1    = dnut_ME,
    dnut_NDF_100 = dnut_NDF / 100,
    dnut_CP_50   = dnut_CP  / 50,
    dnut_EE_10   = dnut_EE  / 10
  ) %>%
  select(B.Code, Species, Product,
         dEI_pct, dY_pct, dCH4_pct,
         dnut_ME_1, dnut_NDF_100, dnut_CP_50, dnut_EE_10)

# 2️⃣ Model-fitting helper -------------------------------------------------
fit_one_milk <- function(sp, n_min = 10) {
  dat <- base_data_milk %>% filter(Species == sp)
  if (nrow(dat) < n_min) return(NULL)

  fit_outcome <- function(y, label) {
    preds <- c("dnut_ME_1","dnut_NDF_100","dnut_CP_50","dnut_EE_10")
    keep  <- complete.cases(dat[, c(y, preds)])
    d     <- dat[keep, , drop = FALSE]
    if (nrow(d) < n_min) return(NULL)

    m <- lm(as.formula(paste0(y, " ~ ", paste(preds, collapse = " + "))), data = d)

    broom::tidy(m, conf.int = TRUE) %>%
      filter(term %in% preds) %>%
      mutate(
        Species = sp,
        Product = "Milk",
        outcome = label,
        term = recode(term,
                      "dnut_ME_1"    = "ΔME (per 1 MJ/kg)",
                      "dnut_NDF_100" = "ΔNDF (per 100 g/kg)",
                      "dnut_CP_50"   = "ΔCP (per 50 g/kg)",
                      "dnut_EE_10"   = "ΔEE (per 10 g/kg)")
      )
  }

  bind_rows(
    fit_outcome("dEI_pct",  "ΔEI (%)"),
    fit_outcome("dY_pct",   "ΔYield (%)"),
    fit_outcome("dCH4_pct", "ΔCH4 (%)")
  )
}

# 3️⃣ Fit models for all milk species -------------------------------------
milk_species <- levels(base_data_milk$Species)
coef_species_milk <- map_dfr(milk_species, fit_one_milk)

if (nrow(coef_species_milk) == 0) {
  stop("No milk species had enough data to fit models.")
}

# 4️⃣ Prepare data ---------------------------------------------------------
coef_species_milk <- coef_species_milk %>%
  mutate(outcome = factor(outcome, levels = c("ΔCH4 (%)", "ΔEI (%)", "ΔYield (%)")))

# Split by species
cattle_df <- coef_species_milk %>% filter(Species == "Cattle")
sheep_df  <- coef_species_milk %>% filter(Species == "Sheep")
goat_df   <- coef_species_milk %>% filter(Species == "Goat")

# 5️⃣ Plot helper ----------------------------------------------------------
# 5️⃣ Plot helper ----------------------------------------------------------
# Use same transparent colors and remove legend
nutrient_colors <- c(
  "ΔCP (per 50 g/kg)"  = "blue4",
  "ΔEE (per 10 g/kg)"  = "darkorange",
  "ΔME (per 1 MJ/kg)"  = "seagreen3",
  "ΔNDF (per 100 g/kg)"= "plum3"
)

plot_species_milk <- function(data, sp_name, show_y = TRUE, show_right = FALSE) {
  ggplot(data,
         aes(x = estimate, y = forcats::fct_rev(term), color = term)) +
    geom_vline(xintercept = 0, linetype = 2, color = "grey70") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0.15, linewidth = 1.1, alpha = 0.5) +  # transparent bars
    geom_point(size = 3, alpha = 0.6) +                          # transparent points
    facet_wrap(
      ~ outcome,
      scales = "free_x",
      ncol = 1,
      strip.position = if (show_right) "right" else "top"
    ) +
    scale_color_manual(values = nutrient_colors) +                # same palette
    labs(
      subtitle = sp_name,
      x = "Estimated effect (95% CI)",
      y = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      strip.text    = element_text(face = "bold", size = 11),
      strip.placement = "outside",
      panel.spacing = unit(1.2, "lines"),
      plot.subtitle = element_text(face = "bold", size = 13, hjust = 0.5),
      axis.text.y   = if (show_y) element_text(face = "bold") else element_blank(),
      axis.ticks.y  = if (show_y) element_line() else element_blank(),
      strip.text.y.right = element_text(face = "bold", angle = -90),
      legend.position = "none"    # remove legend
    )
}


# 6️⃣ Build plots ----------------------------------------------------------
p_cattle <- plot_species_milk(cattle_df, "Cattle (Milk)", show_y = TRUE,  show_right = FALSE)
p_sheep  <- plot_species_milk(sheep_df,  "Sheep (Milk)",  show_y = FALSE, show_right = FALSE)
p_goat   <- plot_species_milk(goat_df,   "Goat (Milk)",   show_y = FALSE, show_right = TRUE)

# 7️⃣ Combine horizontally -------------------------------------------------
final_milk_plot <- p_cattle | p_sheep | p_goat +
  plot_annotation(
    title = "Within-study nutrient shifts vs. ΔEI, ΔYield, and ΔCH₄ (Milk systems)",
    subtitle = "Columns = Species; Rows = Δ outcomes; each facet has its own x-scale",
    caption = "Models use within-study differences; CP per 50 g/kg, NDF per 100 g/kg, EE per 10 g/kg, ME per 1 MJ/kg."
  )

# 8️⃣ Show ----------------------------------------------------------------
final_milk_plot

```

